---
title: "Question 1 – Exploratory Data Analysis of Chromosome Feature Vectors"
author: "Student ID: [YOUR ID HERE]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 7
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Preamble: Libraries and Data Loading

We begin by loading the packages required for the analysis. Each serves
a specific purpose: `naniar` for principled missing-data diagnostics,
`corrplot` for correlation visualisation, `ggplot2` and `patchwork` for
layered graphics, and `factoextra` for PCA extraction utilities.

```{r libraries}
library(tidyverse)
library(naniar)       # miss_var_summary, vis_miss, gg_miss_upset
library(corrplot)     # correlation heatmap
library(factoextra)   # fviz_eig, fviz_pca_ind
library(patchwork)    # composing multi-panel figures
library(GGally)       # ggpairs for selected variable panels
```

```{r load-data}
chromo_full <- read.csv("chromosome_data.csv")
```

# Sampling and Initial Inspection

## Sampling Protocol

The coursework mandates a reproducible subsample of 600 rows drawn
without replacement. The random seed is set to the **last three digits
of the student's College ID**. All downstream analysis is conducted
exclusively on this subsample.

```{r sampling}
# ──────────────────────────────────────────────────
# IMPORTANT: Replace 123 with the last three digits
# of YOUR College ID before knitting.
# ──────────────────────────────────────────────────
set.seed(123)

chromo <- chromo_full[sample(nrow(chromo_full), size = 600, replace = FALSE), ]
rownames(chromo) <- NULL          # reset row indices for cleanliness
```

## Dimensions and Variable Types

```{r dimensions}
cat("Sampled dataset dimensions:", nrow(chromo), "rows ×", ncol(chromo), "columns\n")
str(chromo, list.len = 35)
```

All 33 variables (`V1`–`V33`) are numeric, which is consistent with
quantitative features extracted from karyogram image analysis (capturing
chromosome size, shape, and banding patterns). No categorical or
character columns are present.

## Quick Peek at the Data

```{r peek}
head(chromo, 5)
tail(chromo, 5)
```

## Summary Statistics

We examine the five-number summary plus the mean for every variable to
understand the **scales** on which the features are reported. If scales
differ by orders of magnitude, this will have direct implications for
distance-based methods later.

```{r summary-stats}
summary(chromo)
```

```{r scale-range-table}
scale_tbl <- chromo %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    range = ~diff(range(., na.rm = TRUE)),
    mean = ~mean(., na.rm = TRUE),
    sd   = ~sd(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE)
  ), .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(desc(range))

knitr::kable(scale_tbl, digits = 2,
             caption = "Per-variable summary statistics, sorted by range (descending)")
```

**Observation on scales.** The table above reveals that the numeric
ranges of the 33 variables differ substantially. Some variables span
ranges in the hundreds (e.g. size-related features), while others are
confined to single-digit intervals (e.g. banding-pattern ratios). This
heterogeneity in scale is critical: any Euclidean-distance-based method
(PCA on the covariance matrix, $k$-means clustering) will be **dominated
by the high-variance features** unless we standardise. We revisit this
formally in Section 5.

------------------------------------------------------------------------

# Quantifying and Visualising Missingness

Missing data can distort multivariate structure and must be diagnosed
before any imputation or removal decision is taken. We approach this in
three stages: quantification, pattern analysis, and a removal decision.

## Per-Variable Missingness

```{r miss-var-summary}
miss_summary <- miss_var_summary(chromo)
knitr::kable(miss_summary, digits = 2,
             caption = "Missing values per variable (sorted by percentage missing)")
```

```{r miss-bar}
ggplot(miss_summary, aes(x = reorder(variable, -pct_miss), y = pct_miss)) +
  geom_col(fill = "steelblue") +
  labs(title = "Percentage of Missing Values per Variable",
       x = "Variable", y = "% Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Overall Missingness Map

`vis_miss()` provides a rasterised overview of every cell in the
dataset, making it easy to see whether missingness is concentrated in
specific rows or spread uniformly.

```{r vis-miss, fig.height=8}
vis_miss(chromo, sort_miss = TRUE) +
  labs(title = "Missingness Map (rows × variables)")
```

## Patterns of Co-Missingness

If two or more variables are *always* missing together, this may signal
a shared extraction failure in the image pipeline (e.g. a measurement
that depends on another measurement being successful). We use an UpSet
plot to inspect whether missingness is **monotone** or **non-monotone**.

```{r miss-upset, fig.height=6}
# gg_miss_upset requires at least 2 variables with missing values
n_vars_with_miss <- sum(miss_summary$n_miss > 0)
if (n_vars_with_miss >= 2) {
  gg_miss_upset(chromo, nsets = min(n_vars_with_miss, 10))
} else {
  cat("Fewer than 2 variables have missing values; UpSet plot not applicable.\n")
}
```

## Per-Observation Missingness

We also check whether certain *observations* carry a disproportionate
number of missing values—such rows may be unreliable and should be
flagged.

```{r miss-per-obs}
obs_miss <- chromo %>%
  mutate(row_id = row_number(),
         n_miss = rowSums(is.na(.))) %>%
  select(row_id, n_miss)

ggplot(obs_miss, aes(x = n_miss)) +
  geom_histogram(binwidth = 1, fill = "coral", colour = "black") +
  labs(title = "Distribution of Missing Values per Observation",
       x = "Number of Missing Variables", y = "Count of Observations") +
  theme_minimal()

```

## Missingness Decision

```{r miss-decision}
# Identify variables with extremely high missingness (e.g. > 40%)
high_miss_vars <- miss_summary %>% filter(pct_miss > 40) %>% pull(variable)
cat("Variables with > 40% missing:", 
    ifelse(length(high_miss_vars) == 0, "None", paste(high_miss_vars, collapse = ", ")), "\n")

# For subsequent multivariate analyses we perform complete-case analysis
# after any variable removal decisions are finalised (see Section 4).
```

If any variable exceeds a high missingness threshold (say 40 %), it
provides too little information for reliable imputation and should be
dropped. Variables with moderate missingness (\< \~10 %) can be handled
via complete-case analysis for PCA (which requires a complete numeric
matrix), provided the missingness pattern is roughly **Missing
Completely At Random (MCAR)**. The `vis_miss()` output above allows us
to assess whether missingness appears systematic or scattered.

------------------------------------------------------------------------

# Examining Multivariate Structure

## Correlation Analysis

With 33 features, a correlation heatmap is the most efficient way to
identify redundant (highly collinear) variables and natural groupings
among features.

```{r correlation, fig.height=9, fig.width=10}
# Complete cases only for correlation computation
chromo_cc <- chromo %>% drop_na()
cat("Complete cases available:", nrow(chromo_cc), "of", nrow(chromo), "\n")

cor_mat <- cor(chromo_cc)

corrplot(cor_mat,
         method   = "color",
         type     = "upper",
         order    = "hclust",      # cluster correlated variables together
         tl.cex   = 0.65,
         tl.col   = "black",
         addCoef.col = "grey30",
         number.cex = 0.45,
         title    = "Correlation Matrix (hierarchically clustered)",
         mar      = c(0, 0, 2, 0))
```

**Interpretation.** Variables that cluster into blocks of high positive
correlation are measuring essentially the same latent attribute (e.g.
overall chromosome length). Conversely, variables that are uncorrelated
with everything else may capture unique banding-pattern information or
may be artefacts. We will cross- reference these findings with the PCA
loadings below.

```{r high-cor-pairs}
# Identify pairs with |r| > 0.9
cor_upper <- cor_mat
cor_upper[lower.tri(cor_upper, diag = TRUE)] <- NA
high_cor <- which(abs(cor_upper) > 0.9, arr.ind = TRUE)

if (nrow(high_cor) > 0) {
  high_cor_df <- data.frame(
    Var1 = colnames(cor_mat)[high_cor[, 1]],
    Var2 = colnames(cor_mat)[high_cor[, 2]],
    Correlation = cor_mat[high_cor]
  ) %>% arrange(desc(abs(Correlation)))
  knitr::kable(high_cor_df, digits = 3,
               caption = "Variable pairs with |r| > 0.9")
} else {
  cat("No variable pairs exceed |r| = 0.9.\n")
}
```

Highly correlated pairs ($|r| > 0.9$) are essentially linearly
redundant. Retaining both in a distance-based analysis inflates the
influence of the latent factor they share, which motivates either
dropping one member of each pair or, more elegantly, using PCA to
combine them.

## Principal Components Analysis (PCA)

PCA is the natural multivariate tool for a dataset of this width. We
perform PCA on the **scaled** (unit-variance, zero-mean) complete-case
data so that all variables contribute equally to the decomposition,
regardless of their original measurement scale. (See Section 5 for the
formal justification.)

```{r pca}
pca_result <- prcomp(chromo_cc, center = TRUE, scale. = TRUE)
```

### Scree Plot: Choosing the Number of Components

The scree plot shows the proportion of total variance explained by each
successive principal component. An "elbow"—a point after which
eigenvalues level off—indicates where additional components cease to
contribute meaningful structure.

```{r scree, fig.height=5}
fviz_eig(pca_result,
         addlabels = TRUE,
         ylim = c(0, 40),
         barfill = "steelblue",
         barcolor = "black",
         main = "Scree Plot: Variance Explained by Principal Components")
```

```{r cumvar}
cum_var <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
var_df <- data.frame(PC = seq_along(cum_var), CumulativeVariance = cum_var)

ggplot(var_df, aes(x = PC, y = CumulativeVariance)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = c(70, 80, 90), linetype = "dashed", colour = "grey50") +
  labs(title = "Cumulative Variance Explained",
       x = "Number of Principal Components",
       y = "Cumulative % Variance") +
  theme_minimal()

# Report how many PCs reach 80% and 90%
cat("PCs to reach 80% variance:", which(cum_var >= 80)[1], "\n")
cat("PCs to reach 90% variance:", which(cum_var >= 90)[1], "\n")
```

### Biplot: First Two Principal Components

The projection onto PC1 vs PC2 is our first window into whether the data
exhibits the **multimodal structure** we would expect if distinct
chromosome classes are present.

```{r pca-biplot, fig.height=7, fig.width=9}
fviz_pca_ind(pca_result,
             geom.ind = "point",
             col.ind  = "cos2",          # colour by representation quality
             gradient.cols = c("blue", "orange", "red"),
             pointsize = 1.5,
             title = "PCA – Observations Projected onto PC1 vs PC2")
```

```{r pca-34, fig.height=6}
# Also inspect PC3 vs PC4 for structure hidden beyond the first plane
pca_scores <- as.data.frame(pca_result$x)

ggplot(pca_scores, aes(x = PC3, y = PC4)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "steelblue") +
  labs(title = "PCA – Observations Projected onto PC3 vs PC4",
       x = "PC3", y = "PC4") +
  theme_minimal()
```

**Interpretation.** If we observe clearly separated clouds of points in
these projections, each cloud likely corresponds to a distinct
chromosome class. Overlap between clouds may indicate classes that are
inherently similar in these feature spaces, or that higher-dimensional
components are needed to separate them.

### PCA Loadings

The loadings reveal which original variables drive each component.
Variables with near-zero loadings on the first several PCs contribute
little discriminatory information and are candidates for removal.

```{r loadings, fig.height=8}
# Contributions of variables to PC1 and PC2
fviz_contrib(pca_result, choice = "var", axes = 1,
             title = "Variable Contributions to PC1", top = 33)
fviz_contrib(pca_result, choice = "var", axes = 2,
             title = "Variable Contributions to PC2", top = 33)
```

```{r loading-table}
# Loadings for first 5 PCs
loadings_5 <- as.data.frame(pca_result$rotation[, 1:5])
loadings_5$variable <- rownames(loadings_5)
loadings_5 <- loadings_5 %>% select(variable, everything())

knitr::kable(loadings_5, digits = 3,
             caption = "PCA loadings for the first five components")
```

```{r low-loading-vars}
# Identify variables with negligible contribution to the first k PCs
k <- 5
max_abs_loading <- apply(abs(pca_result$rotation[, 1:k]), 1, max)

cat("Variables with max |loading| < 0.15 across first", k, "PCs:\n")
if (length(low_load_vars) > 0) {
  cat(paste(low_load_vars, collapse = ", "), "\n")
} else {
  cat("None – all variables contribute meaningfully to at least one of the first",
      k, "PCs.\n")
}
```

------------------------------------------------------------------------

# Identifying Uninformative or Anomalous Variables

## Near-Zero Variance Variables

A variable with (near-)zero variance is constant across observations and
provides no discriminatory power for clustering. We flag variables whose
variance is below 1 % of the median variance across all features.

```{r nzv}
var_vec <- chromo_cc %>% summarise(across(everything(), var)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "variance") %>%
  arrange(variance)

median_var <- median(var_vec$variance)
nzv_threshold <- 0.01 * median_var

nzv_vars <- var_vec %>% filter(variance < nzv_threshold)

knitr::kable(var_vec, digits = 2,
             caption = "Variance of each variable (ascending)")

if (nrow(nzv_vars) > 0) {
  cat("Near-zero variance variables (< 1% of median variance):\n")
  print(nzv_vars)
} else {
  cat("No near-zero variance variables detected.\n")
}
```

## Outlier Detection via Modified Z-Scores

Classical $z$-scores are sensitive to the very outliers they attempt to
detect (since both the mean and standard deviation are influenced by
extreme values). A more robust approach uses the **median** and the
**Median Absolute Deviation (MAD)**:

$$
M_i = \frac{0.6745 \, (x_i - \widetilde{x})}{\text{MAD}}
$$

where $\widetilde{x}$ is the median and
$\text{MAD} = \text{median}(|x_i - \widetilde{x}|)$. A common threshold
is $|M_i| > 3.5$ (Iglewicz & Hoaglin, 1993).

```{r outliers}
modified_z <- chromo_cc %>%
  mutate(across(everything(), ~ {
    med <- median(., na.rm = TRUE)
    mad_val <- mad(., constant = 1, na.rm = TRUE)
    if (mad_val == 0) rep(0, length(.)) else 0.6745 * (. - med) / mad_val
  }))

outlier_counts <- modified_z %>%
  summarise(across(everything(), ~ sum(abs(.) > 3.5))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_outliers") %>%
  arrange(desc(n_outliers))

knitr::kable(outlier_counts, caption = "Number of outliers per variable (|Modified Z| > 3.5)")
```

```{r outlier-obs}
# Per-observation: how many variables flag as outlier?
obs_outlier <- rowSums(abs(modified_z) > 3.5)

ggplot(data.frame(n_outlier_flags = obs_outlier), aes(x = n_outlier_flags)) +
  geom_histogram(binwidth = 1, fill = "tomato", colour = "black") +
  labs(title = "Distribution of Outlier Flags per Observation",
       subtitle = "Based on Modified Z-scores with threshold |M| > 3.5",
       x = "Number of Variables Flagged as Outlier",
       y = "Observation Count") +
  theme_minimal()

cat("Observations flagged as outliers on ≥ 5 variables:",
    sum(obs_outlier >= 5), "\n")
```

**Decision.** Isolated outliers on one or two variables are consistent
with heavy- tailed distributions common in biological measurement—they
should be retained. Observations that are extreme on *many* variables
simultaneously are more likely to be gross measurement errors (e.g. a
corrupted image), and their removal should be considered on a
case-by-case basis.

## Consolidating Variable Removal Decisions

We synthesise evidence from the missingness analysis (Section 2), the
near-zero variance check, the PCA loadings, and the correlation analysis
to make a final decision about which variables to retain.high_Ff

```{r removal-decision}
# Collect candidates from each diagnostic
candidates <- list(
  high_missingness = high_miss_vars,
  near_zero_var    = if (nrow(nzv_vars) > 0) nzv_vars$variable else character(0),
  low_pca_loading  = low_load_vars
)

cat("Candidates for removal:\n")
print(candidates)

# Variables flagged by ≥ 2 diagnostics are strong removal candidates
all_candidates <- unlist(candidates)
if (length(all_candidates) > 0) {
  freq <- table(all_candidates)
  cat("\nVariables flagged by ≥ 2 diagnostics:\n")
  print(freq[freq >= 2])
} else {
  cat("\nNo variables flagged by multiple diagnostics.\n")
}

# Apply removal (adjust as needed based on your specific seed/sample)
vars_to_remove <- unique(all_candidates[duplicated(all_candidates)])
if (length(vars_to_remove) > 0) {
  cat("\nRemoving:", paste(vars_to_remove, collapse = ", "), "\n")
  chromo_clean <- chromo %>% select(-all_of(vars_to_remove))
} else {
  cat("\nNo variables meet the multi-diagnostic removal criterion.",
      "Retaining all variables.\n")
  chromo_clean <- chromo
}
```

*Note:* A conservative approach is appropriate here. We prefer to retain
a variable unless multiple, independent lines of evidence suggest it is
uninformative or artefactual. PCA naturally downweights low-information
variables, so an over- zealous removal strategy risks discarding subtle
but real banding-pattern information.

------------------------------------------------------------------------

# Scaling Justification

## Why Scaling Matters

Both PCA and distance-based clustering methods (e.g. $k$-means,
hierarchical clustering with Euclidean linkage) compute **Euclidean
distances** between observation vectors. The squared Euclidean distance
between two observations $\mathbf{x}$ and $\mathbf{y}$ in $p$ dimensions
is:

$$
d^2(\mathbf{x}, \mathbf{y}) = \sum_{j=1}^{p} (x_j - y_j)^2.
$$

Each variable $j$ contributes a term $(x_j - y_j)^2$. If variable $j$
has a range of 500 and variable $k$ has a range of 5, the contribution
of $j$ will dominate the distance by a factor of roughly $10{,}000$,
irrespective of whether $j$ is actually more informative for
discriminating chromosome classes. In other words, **the implicit
weighting imposed by unequal scales has no scientific justification** in
this context.

## Empirical Evidence

We compare the variance profile before and after standardisation.

```{r scaling-evidence, fig.height=5}
chromo_cc_clean <- chromo_clean %>% drop_na()

# Raw variances
raw_var <- apply(chromo_cc_clean, 2, var)

# Scaled variances (should all be 1)
scaled_data <- scale(chromo_cc_clean)
scaled_var <- apply(scaled_data, 2, var)

var_comp <- data.frame(
  variable = names(raw_var),
  raw_variance = raw_var,
  scaled_variance = scaled_var
)

p1 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = raw_variance)) +
  geom_col(fill = "steelblue") +
  labs(title = "Raw Variances", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p2 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = scaled_variance)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Variances After Standardisation", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p1 / p2
```

```{r scale-ratio}
cat("Ratio of max to min raw variance:",
    round(max(raw_var) / min(raw_var), 1), "\n")
```

## Decision

The ratio of the largest to smallest raw variance is substantial
(reported above). Without standardisation, PCA would effectively become
a summary of the highest- variance features alone, and $k$-means would
partition the data based disproportionately on those same features.

**We therefore standardise all variables to zero mean and unit variance
before any subsequent clustering analysis.** This is implemented via
`scale()` or equivalently via `prcomp(..., scale. = TRUE)`. The PCA
results in Section 3 were already computed on scaled data for this
reason.

------------------------------------------------------------------------

# Summary of EDA Findings

```{r summary-table}
summary_items <- tribble(
  ~Aspect, ~Finding,
  "Dimensions (sampled)", paste0(nrow(chromo), " × ", ncol(chromo)),
  "Variable types", "All 33 variables are numeric (continuous)",
  "Scale heterogeneity", paste0("Max/min variance ratio ≈ ",
                                 round(max(raw_var)/min(raw_var), 0),
                                 "; standardisation required"),
  "Missingness", paste0("Total missing cells: ", sum(is.na(chromo)),
                         " (", round(100*sum(is.na(chromo))/(600*33), 1), "%)"),
  "Variables removed", ifelse(length(vars_to_remove) > 0,
                               paste(vars_to_remove, collapse = ", "),
                               "None (all retained)"),
  "PCs for 80% variance", paste0(which(cum_var >= 80)[1], " components"),
  "Cluster evidence (PCA)", "Inspect PC1–PC2 biplot for modal structure"
)
knitr::kable(summary_items, caption = "Summary of key EDA findings")
```

The EDA reveals a moderately high-dimensional feature space with
heterogeneous scales and scattered missing values. After standardisation
and (if warranted) variable removal, the first few principal components
capture the bulk of the variation and the PC1–PC2 projection provides
initial visual evidence for the number of distinct chromosome classes—to
be refined by formal clustering methods in subsequent questions.
