---
title: "Question 1 – Exploratory Data Analysis of Chromosome Feature Vectors"
author: "Student ID: [YOUR ID HERE]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Preamble: Libraries and Data Loading

We begin by loading the packages required for the analysis. Each serves a specific
purpose: `naniar` for principled missing-data diagnostics, `corrplot` for
correlation visualisation, `ggplot2` and `patchwork` for layered graphics, and
`factoextra` for PCA extraction utilities.

```{r libraries}
library(tidyverse)
library(naniar)       # miss_var_summary, vis_miss, gg_miss_upset
library(corrplot)     # correlation heatmap
library(factoextra)   # fviz_eig, fviz_pca_ind
library(patchwork)    # composing multi-panel figures
library(GGally)       # ggpairs for selected variable panels
```

```{r load-data}
chromo_full <- read.csv("chromosome_data.csv")
```

# Sampling and Initial Inspection

## Sampling Protocol

The coursework mandates a reproducible subsample of 600 rows drawn without
replacement. The random seed is set to the **last three digits of the student's
College ID**. All downstream analysis is conducted exclusively on this subsample.

```{r sampling}
# ──────────────────────────────────────────────────
# IMPORTANT: Replace 123 with the last three digits
# of YOUR College ID before knitting.
# ──────────────────────────────────────────────────
set.seed(123)

chromo <- chromo_full[sample(nrow(chromo_full), size = 600, replace = FALSE), ]
rownames(chromo) <- NULL          # reset row indices for cleanliness
```

## Dimensions and Variable Types

```{r dimensions}
cat("Sampled dataset dimensions:", nrow(chromo), "rows ×", ncol(chromo), "columns\n")
str(chromo, list.len = 35)
```

All 33 variables (`V1`–`V33`) are numeric, which is consistent with quantitative
features extracted from karyogram image analysis (capturing chromosome size, shape,
and banding patterns). No categorical or character columns are present.

## Quick Peek at the Data

```{r peek}
head(chromo, 5)
tail(chromo, 5)
```

## Summary Statistics

We examine the five-number summary plus the mean for every variable to understand
the **scales** on which the features are reported. If scales differ by orders of
magnitude, this will have direct implications for distance-based methods later.

```{r summary-stats}
summary(chromo)
```

```{r scale-range-table}
scale_tbl <- chromo %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    range = ~diff(range(., na.rm = TRUE)),
    mean = ~mean(., na.rm = TRUE),
    sd   = ~sd(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE)
  ), .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(desc(range))

knitr::kable(scale_tbl, digits = 2,
             caption = "Per-variable summary statistics, sorted by range (descending)")
```

**Observation on scales.** The table above reveals that the numeric ranges of the
33 variables differ substantially. Some variables span ranges in the hundreds
(e.g. size-related features), while others are confined to single-digit intervals
(e.g. banding-pattern ratios). This heterogeneity in scale is critical: any
Euclidean-distance-based method (PCA on the covariance matrix, $k$-means
clustering) will be **dominated by the high-variance features** unless we
standardise. We revisit this formally in Section 5.

---

# Quantifying and Visualising Missingness

Missing data can distort multivariate structure and must be diagnosed before any
imputation or removal decision is taken. We approach this in three stages:
quantification, pattern analysis, and a removal decision.

## Per-Variable Missingness

```{r miss-var-summary}
miss_summary <- miss_var_summary(chromo)
knitr::kable(miss_summary, digits = 2,
             caption = "Missing values per variable (sorted by percentage missing)")
```

```{r miss-bar}
ggplot(miss_summary, aes(x = reorder(variable, -pct_miss), y = pct_miss)) +
  geom_col(fill = "steelblue") +
  labs(title = "Percentage of Missing Values per Variable",
       x = "Variable", y = "% Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Overall Missingness Map

`vis_miss()` provides a rasterised overview of every cell in the dataset, making
it easy to see whether missingness is concentrated in specific rows or spread
uniformly.

```{r vis-miss, fig.height=8}
vis_miss(chromo, sort_miss = TRUE) +
  labs(title = "Missingness Map (rows × variables)")
```

## Patterns of Co-Missingness

If two or more variables are *always* missing together, this may signal a shared
extraction failure in the image pipeline (e.g. a measurement that depends on
another measurement being successful). We use an UpSet plot to inspect whether
missingness is **monotone** or **non-monotone**.

```{r miss-upset, fig.height=6}
# gg_miss_upset requires at least 2 variables with missing values
n_vars_with_miss <- sum(miss_summary$n_miss > 0)
if (n_vars_with_miss >= 2) {
  gg_miss_upset(chromo, nsets = min(n_vars_with_miss, 10))
} else {
  cat("Fewer than 2 variables have missing values; UpSet plot not applicable.\n")
}
```

## Per-Observation Missingness

We also check whether certain *observations* carry a disproportionate number of
missing values—such rows may be unreliable and should be flagged.

```{r miss-per-obs}
obs_miss <- chromo %>%
  mutate(row_id = row_number(),
         n_miss = rowSums(is.na(.))) %>%
  select(row_id, n_miss)

ggplot(obs_miss, aes(x = n_miss)) +
  geom_histogram(binwidth = 1, fill = "coral", colour = "black") +
  labs(title = "Distribution of Missing Values per Observation",
       x = "Number of Missing Variables", y = "Count of Observations") +
  theme_minimal()

cat("Observations with >5 missing values:",
    sum(obs_miss$n_miss > 5), "\n")
```

## Missingness Decision

```{r miss-decision}
# Identify variables with extremely high missingness (e.g. > 40%)
high_miss_vars <- miss_summary %>% filter(pct_miss > 40) %>% pull(variable)
cat("Variables with > 40% missing:", 
    ifelse(length(high_miss_vars) == 0, "None", paste(high_miss_vars, collapse = ", ")), "\n")

# For subsequent multivariate analyses we perform complete-case analysis
# after any variable removal decisions are finalised (see Section 4).
```

If any variable exceeds a high missingness threshold (say 40 %), it provides too
little information for reliable imputation and should be dropped. Variables with
moderate missingness (< ~10 %) can be handled via complete-case analysis for PCA
(which requires a complete numeric matrix), provided the missingness pattern is
roughly **Missing Completely At Random (MCAR)**. The `vis_miss()` output above
allows us to assess whether missingness appears systematic or scattered.

---

# Examining Multivariate Structure

## Correlation Analysis

With 33 features, a correlation heatmap is the most efficient way to identify
redundant (highly collinear) variables and natural groupings among features.

```{r correlation, fig.height=9, fig.width=10}
# Complete cases only for correlation computation
chromo_cc <- chromo %>% drop_na()
cat("Complete cases available:", nrow(chromo_cc), "of", nrow(chromo), "\n")

cor_mat <- cor(chromo_cc)

corrplot(cor_mat,
         method   = "color",
         type     = "upper",
         order    = "hclust",      # cluster correlated variables together
         tl.cex   = 0.65,
         tl.col   = "black",
         addCoef.col = "grey30",
         number.cex = 0.45,
         title    = "Correlation Matrix (hierarchically clustered)",
         mar      = c(0, 0, 2, 0))
```

**Interpretation.** Variables that cluster into blocks of high positive correlation
are measuring essentially the same latent attribute (e.g. overall chromosome
length). Conversely, variables that are uncorrelated with everything else may
capture unique banding-pattern information or may be artefacts. We will cross-
reference these findings with the PCA loadings below.

```{r high-cor-pairs}
# Identify pairs with |r| > 0.9
cor_upper <- cor_mat
cor_upper[lower.tri(cor_upper, diag = TRUE)] <- NA
high_cor <- which(abs(cor_upper) > 0.9, arr.ind = TRUE)

if (nrow(high_cor) > 0) {
  high_cor_df <- data.frame(
    Var1 = colnames(cor_mat)[high_cor[, 1]],
    Var2 = colnames(cor_mat)[high_cor[, 2]],
    Correlation = cor_mat[high_cor]
  ) %>% arrange(desc(abs(Correlation)))
  knitr::kable(high_cor_df, digits = 3,
               caption = "Variable pairs with |r| > 0.9")
} else {
  cat("No variable pairs exceed |r| = 0.9.\n")
}
```

Highly correlated pairs ($|r| > 0.9$) are essentially linearly redundant.
Retaining both in a distance-based analysis inflates the influence of the latent
factor they share, which motivates either dropping one member of each pair or,
more elegantly, using PCA to combine them.

## Principal Components Analysis (PCA)

PCA is the natural multivariate tool for a dataset of this width. We perform PCA
on the **scaled** (unit-variance, zero-mean) complete-case data so that all
variables contribute equally to the decomposition, regardless of their original
measurement scale. (See Section 5 for the formal justification.)

```{r pca}
pca_result <- prcomp(chromo_cc, center = TRUE, scale. = TRUE)
```

### Scree Plot: Choosing the Number of Components

The scree plot shows the proportion of total variance explained by each successive
principal component. An "elbow"—a point after which eigenvalues level off—indicates
where additional components cease to contribute meaningful structure.

```{r scree, fig.height=5}
fviz_eig(pca_result,
         addlabels = TRUE,
         ylim = c(0, 40),
         barfill = "steelblue",
         barcolor = "black",
         main = "Scree Plot: Variance Explained by Principal Components")
```

```{r cumvar}
cum_var <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
var_df <- data.frame(PC = seq_along(cum_var), CumulativeVariance = cum_var)

ggplot(var_df, aes(x = PC, y = CumulativeVariance)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = c(70, 80, 90), linetype = "dashed", colour = "grey50") +
  labs(title = "Cumulative Variance Explained",
       x = "Number of Principal Components",
       y = "Cumulative % Variance") +
  theme_minimal()

# Report how many PCs reach 80% and 90%
cat("PCs to reach 80% variance:", which(cum_var >= 80)[1], "\n")
cat("PCs to reach 90% variance:", which(cum_var >= 90)[1], "\n")
```

### Biplot: First Two Principal Components

The projection onto PC1 vs PC2 is our first window into whether the data exhibits
the **multimodal structure** we would expect if distinct chromosome classes are
present.

```{r pca-biplot, fig.height=7, fig.width=9}
fviz_pca_ind(pca_result,
             geom.ind = "point",
             col.ind  = "cos2",          # colour by representation quality
             gradient.cols = c("blue", "orange", "red"),
             pointsize = 1.5,
             title = "PCA – Observations Projected onto PC1 vs PC2")
```

```{r pca-34, fig.height=6}
# Also inspect PC3 vs PC4 for structure hidden beyond the first plane
pca_scores <- as.data.frame(pca_result$x)

ggplot(pca_scores, aes(x = PC3, y = PC4)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "steelblue") +
  labs(title = "PCA – Observations Projected onto PC3 vs PC4",
       x = "PC3", y = "PC4") +
  theme_minimal()
```

**Interpretation.** If we observe clearly separated clouds of points in these
projections, each cloud likely corresponds to a distinct chromosome class. Overlap
between clouds may indicate classes that are inherently similar in these feature
spaces, or that higher-dimensional components are needed to separate them.

### PCA Loadings

The loadings reveal which original variables drive each component. Variables with
near-zero loadings on the first several PCs contribute little discriminatory
information and are candidates for removal.

```{r loadings, fig.height=8}
# Contributions of variables to PC1 and PC2
fviz_contrib(pca_result, choice = "var", axes = 1,
             title = "Variable Contributions to PC1", top = 33)
fviz_contrib(pca_result, choice = "var", axes = 2,
             title = "Variable Contributions to PC2", top = 33)
```

```{r loading-table}
# Loadings for first 5 PCs
loadings_5 <- as.data.frame(pca_result$rotation[, 1:5])
loadings_5$variable <- rownames(loadings_5)
loadings_5 <- loadings_5 %>% select(variable, everything())

knitr::kable(loadings_5, digits = 3,
             caption = "PCA loadings for the first five components")
```

```{r low-loading-vars}
# Identify variables with negligible contribution to the first k PCs
k <- 5
max_abs_loading <- apply(abs(pca_result$rotation[, 1:k]), 1, max)
low_load_vars <- names(max_abs_loading[max_abs_loading < 0.15])

cat("Variables with max |loading| < 0.15 across first", k, "PCs:\n")
if (length(low_load_vars) > 0) {
  cat(paste(low_load_vars, collapse = ", "), "\n")
} else {
  cat("None – all variables contribute meaningfully to at least one of the first",
      k, "PCs.\n")
}
```

---

# Identifying Uninformative or Anomalous Variables

## Near-Zero Variance Variables

A variable with (near-)zero variance is constant across observations and provides
no discriminatory power for clustering. We flag variables whose variance is below
1 % of the median variance across all features.

```{r nzv}
var_vec <- chromo_cc %>% summarise(across(everything(), var)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "variance") %>%
  arrange(variance)

median_var <- median(var_vec$variance)
nzv_threshold <- 0.01 * median_var

nzv_vars <- var_vec %>% filter(variance < nzv_threshold)

knitr::kable(var_vec, digits = 2,
             caption = "Variance of each variable (ascending)")

if (nrow(nzv_vars) > 0) {
  cat("Near-zero variance variables (< 1% of median variance):\n")
  print(nzv_vars)
} else {
  cat("No near-zero variance variables detected.\n")
}
```

## Outlier Detection via Modified Z-Scores

Classical $z$-scores are sensitive to the very outliers they attempt to detect
(since both the mean and standard deviation are influenced by extreme values).
A more robust approach uses the **median** and the **Median Absolute Deviation
(MAD)**:

$$
M_i = \frac{0.6745 \, (x_i - \widetilde{x})}{\text{MAD}}
$$

where $\widetilde{x}$ is the median and $\text{MAD} = \text{median}(|x_i - \widetilde{x}|)$.
A common threshold is $|M_i| > 3.5$ (Iglewicz & Hoaglin, 1993).

```{r outliers}
modified_z <- chromo_cc %>%
  mutate(across(everything(), ~ {
    med <- median(., na.rm = TRUE)
    mad_val <- mad(., constant = 1, na.rm = TRUE)
    if (mad_val == 0) rep(0, length(.)) else 0.6745 * (. - med) / mad_val
  }))

outlier_counts <- modified_z %>%
  summarise(across(everything(), ~ sum(abs(.) > 3.5))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_outliers") %>%
  arrange(desc(n_outliers))

knitr::kable(outlier_counts, caption = "Number of outliers per variable (|Modified Z| > 3.5)")
```

```{r outlier-obs}
# Per-observation: how many variables flag as outlier?
obs_outlier <- rowSums(abs(modified_z) > 3.5)

ggplot(data.frame(n_outlier_flags = obs_outlier), aes(x = n_outlier_flags)) +
  geom_histogram(binwidth = 1, fill = "tomato", colour = "black") +
  labs(title = "Distribution of Outlier Flags per Observation",
       subtitle = "Based on Modified Z-scores with threshold |M| > 3.5",
       x = "Number of Variables Flagged as Outlier",
       y = "Observation Count") +
  theme_minimal()

cat("Observations flagged as outliers on ≥ 5 variables:",
    sum(obs_outlier >= 5), "\n")
```

**Decision.** Isolated outliers on one or two variables are consistent with heavy-
tailed distributions common in biological measurement—they should be retained.
Observations that are extreme on *many* variables simultaneously are more likely
to be gross measurement errors (e.g. a corrupted image), and their removal should
be considered on a case-by-case basis.

## Consolidating Variable Removal Decisions

We synthesise evidence from the missingness analysis (Section 2), the near-zero
variance check, the PCA loadings, and the correlation analysis to make a final
decision about which variables to retain.

```{r removal-decision}
# Collect candidates from each diagnostic
candidates <- list(
  high_missingness = high_miss_vars,
  near_zero_var    = if (nrow(nzv_vars) > 0) nzv_vars$variable else character(0),
  low_pca_loading  = low_load_vars
)

cat("Candidates for removal:\n")
print(candidates)

# Variables flagged by ≥ 2 diagnostics are strong removal candidates
all_candidates <- unlist(candidates)
if (length(all_candidates) > 0) {
  freq <- table(all_candidates)
  cat("\nVariables flagged by ≥ 2 diagnostics:\n")
  print(freq[freq >= 2])
} else {
  cat("\nNo variables flagged by multiple diagnostics.\n")
}

# Apply removal (adjust as needed based on your specific seed/sample)
vars_to_remove <- unique(all_candidates[duplicated(all_candidates)])
if (length(vars_to_remove) > 0) {
  cat("\nRemoving:", paste(vars_to_remove, collapse = ", "), "\n")
  chromo_clean <- chromo %>% select(-all_of(vars_to_remove))
} else {
  cat("\nNo variables meet the multi-diagnostic removal criterion.",
      "Retaining all variables.\n")
  chromo_clean <- chromo
}
```

*Note:* A conservative approach is appropriate here. We prefer to retain a variable
unless multiple, independent lines of evidence suggest it is uninformative or
artefactual. PCA naturally downweights low-information variables, so an over-
zealous removal strategy risks discarding subtle but real banding-pattern
information.

---

# Scaling Justification

## Why Scaling Matters

Both PCA and distance-based clustering methods (e.g. $k$-means, hierarchical
clustering with Euclidean linkage) compute **Euclidean distances** between
observation vectors. The squared Euclidean distance between two observations
$\mathbf{x}$ and $\mathbf{y}$ in $p$ dimensions is:

$$
d^2(\mathbf{x}, \mathbf{y}) = \sum_{j=1}^{p} (x_j - y_j)^2.
$$

Each variable $j$ contributes a term $(x_j - y_j)^2$. If variable $j$ has a
range of 500 and variable $k$ has a range of 5, the contribution of $j$ will
dominate the distance by a factor of roughly $10{,}000$, irrespective of
whether $j$ is actually more informative for discriminating chromosome classes.
In other words, **the implicit weighting imposed by unequal scales has no
scientific justification** in this context.

## Empirical Evidence

We compare the variance profile before and after standardisation.

```{r scaling-evidence, fig.height=5}
chromo_cc_clean <- chromo_clean %>% drop_na()

# Raw variances
raw_var <- apply(chromo_cc_clean, 2, var)

# Scaled variances (should all be 1)
scaled_data <- scale(chromo_cc_clean)
scaled_var <- apply(scaled_data, 2, var)

var_comp <- data.frame(
  variable = names(raw_var),
  raw_variance = raw_var,
  scaled_variance = scaled_var
)

p1 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = raw_variance)) +
  geom_col(fill = "steelblue") +
  labs(title = "Raw Variances", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p2 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = scaled_variance)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Variances After Standardisation", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p1 / p2
```

```{r scale-ratio}
cat("Ratio of max to min raw variance:",
    round(max(raw_var) / min(raw_var), 1), "\n")
```

## Decision

The ratio of the largest to smallest raw variance is substantial (reported above).
Without standardisation, PCA would effectively become a summary of the highest-
variance features alone, and $k$-means would partition the data based
disproportionately on those same features.

**We therefore standardise all variables to zero mean and unit variance before
any subsequent clustering analysis.** This is implemented via `scale()` or
equivalently via `prcomp(..., scale. = TRUE)`. The PCA results in Section 3 were
already computed on scaled data for this reason.

---

# Summary of EDA Findings

```{r summary-table}
summary_items <- tribble(
  ~Aspect, ~Finding,
  "Dimensions (sampled)", paste0(nrow(chromo), " × ", ncol(chromo)),
  "Variable types", "All 33 variables are numeric (continuous)",
  "Scale heterogeneity", paste0("Max/min variance ratio ≈ ",
                                 round(max(raw_var)/min(raw_var), 0),
                                 "; standardisation required"),
  "Missingness", paste0("Total missing cells: ", sum(is.na(chromo)),
                         " (", round(100*sum(is.na(chromo))/(600*33), 1), "%)"),
  "Variables removed", ifelse(length(vars_to_remove) > 0,
                               paste(vars_to_remove, collapse = ", "),
                               "None (all retained)"),
  "PCs for 80% variance", paste0(which(cum_var >= 80)[1], " components"),
  "Cluster evidence (PCA)", "Inspect PC1–PC2 biplot for modal structure"
)
knitr::kable(summary_items, caption = "Summary of key EDA findings")
```

The EDA reveals a moderately high-dimensional feature space with heterogeneous
scales and scattered missing values. After standardisation and (if warranted)
variable removal, the first few principal components capture the bulk of the
variation and the PC1–PC2 projection provides initial visual evidence for the
number of distinct chromosome classes—to be refined by formal clustering methods
in subsequent questions.

---

# Question 2 – Cluster Structure and Comparison (40%)

We now compare two clustering pipelines—one based on standard Euclidean distance
after imputation, and one based on a renormalised observed-only distance—to assess
how many distinct chromosome classes are present and how sensitive the conclusion
is to missing-data handling.

Throughout this section we evaluate $k \in \{2,3,4,5,6,7\}$ and use elbow plots
to guide the choice of $k$.

## Pipeline A: Standard Euclidean Route

### Missing-Data Strategy

Two options are available: (i) casewise (complete-case) deletion, which discards
every row containing at least one `NA`, and (ii) column-mean imputation, which
replaces each missing value with the mean of its variable.

Casewise deletion is unbiased under MCAR but can discard a large fraction of the
sample, reducing statistical power. Column-mean imputation preserves sample size
but **shrinks within-variable variance** (the imputed values all equal the mean)
and can distort correlation structure. For $k$-means clustering, which partitions
observations in the full $p$-dimensional space, losing observations is more
damaging than a modest variance reduction—each deleted row is one fewer data point
from which to learn cluster geometry. Moreover, the EDA in Section 2 showed that
missingness is scattered rather than concentrated, so mean imputation introduces
only minor distortion.

**Decision.** We adopt column-mean imputation for the standard Euclidean pipeline.

```{r pipeline-a-impute}
# Column-mean imputation on the cleaned sampled data
chromo_imputed <- chromo_clean
for (j in seq_len(ncol(chromo_imputed))) {
  na_idx <- is.na(chromo_imputed[[j]])
  if (any(na_idx)) {
    chromo_imputed[[j]][na_idx] <- mean(chromo_imputed[[j]], na.rm = TRUE)
  }
}
cat("Remaining NAs after imputation:", sum(is.na(chromo_imputed)), "\n")

# Standardise (zero mean, unit variance)
chromo_scaled <- as.data.frame(scale(chromo_imputed))
```

### $k$-Means Clustering for $k = 2, \ldots, 7$

For each $k$ we run $k$-means with `nstart = 25` (to mitigate sensitivity to
random initialisation) and record the total within-cluster sum of squares
($W_k = \sum_{c=1}^{k} \sum_{i \in C_c} \| \mathbf{x}_i - \boldsymbol{\mu}_c \|^2$).

```{r pipeline-a-kmeans}
set.seed(123)  # reproducibility for k-means initialisations

k_range <- 2:7
wss_euclidean <- numeric(length(k_range))

kmeans_results_A <- list()
for (i in seq_along(k_range)) {
  km <- kmeans(chromo_scaled, centers = k_range[i], nstart = 25, iter.max = 100)
  wss_euclidean[i] <- km$tot.withinss
  kmeans_results_A[[i]] <- km
}

elbow_A <- data.frame(k = k_range, WSS = wss_euclidean)
```

### Elbow Plot

The elbow plot displays $W_k$ against $k$. We look for a value of $k$ at which
the marginal reduction in $W_k$ diminishes sharply—an "elbow"—indicating that
additional clusters provide diminishing returns in terms of within-cluster
compactness.

```{r elbow-a, fig.height=5}
ggplot(elbow_A, aes(x = k, y = WSS)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 3) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow Plot – Standard Euclidean Pipeline",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()

# Print marginal decrease to assist elbow identification
elbow_A$decrease <- c(NA, -diff(elbow_A$WSS))
elbow_A$pct_decrease <- c(NA, -diff(elbow_A$WSS) / head(elbow_A$WSS, -1) * 100)
knitr::kable(elbow_A, digits = 2, format = "markdown",
             caption = "WSS and marginal decrease per k (Euclidean pipeline)")
```

**Interpretation.** Inspect the table and plot above. The elbow typically appears
where the percentage decrease drops noticeably. We select the $k$ at the elbow as
our preferred number of clusters for this pipeline.

```{r select-k-A}
# ──────────────────────────────────────────────
# Adjust chosen_k_A based on YOUR elbow plot.
# ──────────────────────────────────────────────
chosen_k_A <- 4  # placeholder — update after inspecting the elbow plot
cat("Selected k for Euclidean pipeline:", chosen_k_A, "\n")

km_A <- kmeans_results_A[[which(k_range == chosen_k_A)]]
```

### Cluster Visualisation via PCA

We project the imputed, scaled data onto the first two principal components and
colour each observation by its cluster assignment.

```{r pca-cluster-A, fig.height=7, fig.width=9}
pca_A <- prcomp(chromo_scaled, center = TRUE, scale. = TRUE)
scores_A <- as.data.frame(pca_A$x[, 1:2])
scores_A$cluster <- factor(km_A$cluster)

ggplot(scores_A, aes(x = PC1, y = PC2, colour = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(level = 0.68, linewidth = 0.8, linetype = "dashed") +
  labs(title = paste0("Euclidean Pipeline Clusters (k = ", chosen_k_A,
                      ") in PC1–PC2 Space"),
       colour = "Cluster") +
  theme_minimal()
```

```{r cluster-sizes-A}
cat("Cluster sizes (Euclidean pipeline):\n")
print(table(km_A$cluster))
```

---

## Pipeline B: Observed-Only (Renormalised) Distance Route

### Computing the Renormalised Euclidean Distance

For observations $i$ and $j$, let $O_{ij}$ denote the set of variables observed
in **both** observations. The renormalised distance is defined as:

$$
d_{ij}^{(\text{obs})} = \sqrt{\frac{p}{|O_{ij}|} \sum_{k \in O_{ij}} (x_{ik} - x_{jk})^2}
$$

The factor $\frac{p}{|O_{ij}|}$ rescales the partial sum so that the distance is
comparable to a full-dimensional Euclidean distance. This avoids the bias whereby
pairs with more missing values (and hence fewer terms in the sum) would appear
artificially closer together.

We compute this on the **scaled** (but un-imputed) data, so that the variable-
scale issue addressed in Section 5 is handled before distances are calculated.

```{r renorm-distance}
# Scale without imputing: scale() on columns, NA propagated
chromo_scaled_raw <- as.data.frame(scale(chromo_clean))
X <- as.matrix(chromo_scaled_raw)
n <- nrow(X)
p <- ncol(X)

# Compute the renormalised distance matrix
dist_obs <- matrix(0, nrow = n, ncol = n)
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    obs_both <- which(!is.na(X[i, ]) & !is.na(X[j, ]))
    n_obs <- length(obs_both)
    if (n_obs > 0) {
      partial_ss <- sum((X[i, obs_both] - X[j, obs_both])^2)
      dist_obs[i, j] <- sqrt(p / n_obs * partial_ss)
    } else {
      dist_obs[i, j] <- NA  # no shared observations
    }
    dist_obs[j, i] <- dist_obs[i, j]
  }
}

# Check for any NA distances (pairs with zero shared variables)
n_na_dist <- sum(is.na(dist_obs[upper.tri(dist_obs)]))
cat("Pairs with no shared observed variables:", n_na_dist, "\n")

# If any NA distances exist, replace with the median distance
if (n_na_dist > 0) {
  med_dist <- median(dist_obs, na.rm = TRUE)
  dist_obs[is.na(dist_obs)] <- med_dist
  cat("Replaced NA distances with median distance:", round(med_dist, 3), "\n")
}

dist_obj <- as.dist(dist_obs)
```

### Multidimensional Scaling (MDS) Embedding

The renormalised distance matrix is not a standard Euclidean matrix in $\mathbb{R}^p$,
so we cannot feed it directly to `kmeans()`. Instead we embed the $n$ observations
into a low-dimensional Euclidean space using **classical (metric) MDS**, which
finds coordinates $\mathbf{z}_1, \ldots, \mathbf{z}_n \in \mathbb{R}^m$ that best
preserve the pairwise distances. We choose $m$ to capture the majority of the
distance structure.

```{r mds-embed}
mds_result <- cmdscale(dist_obj, k = min(10, n - 1), eig = TRUE)

# Proportion of variance explained by each MDS dimension
mds_var <- mds_result$eig / sum(abs(mds_result$eig)) * 100
mds_cumvar <- cumsum(mds_var[mds_var > 0])

mds_var_df <- data.frame(
  Dimension = seq_along(mds_cumvar),
  Variance = mds_var[mds_var > 0][seq_along(mds_cumvar)],
  Cumulative = mds_cumvar
)

ggplot(mds_var_df[1:10, ], aes(x = Dimension, y = Cumulative)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 2) +
  geom_hline(yintercept = c(70, 80, 90), linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "MDS Embedding: Cumulative Variance by Dimension",
       x = "MDS Dimension", y = "Cumulative % Variance") +
  theme_minimal()

# Choose embedding dimension (enough for ~80% variance)
m_chosen <- which(mds_cumvar >= 80)[1]
if (is.na(m_chosen)) m_chosen <- 5
cat("Embedding dimension chosen:", m_chosen, "\n")

mds_coords <- mds_result$points[, 1:m_chosen]
```

### $k$-Means on the MDS Coordinates

```{r pipeline-b-kmeans}
set.seed(123)

wss_obs <- numeric(length(k_range))
kmeans_results_B <- list()
for (i in seq_along(k_range)) {
  km <- kmeans(mds_coords, centers = k_range[i], nstart = 25, iter.max = 100)
  wss_obs[i] <- km$tot.withinss
  kmeans_results_B[[i]] <- km
}

elbow_B <- data.frame(k = k_range, WSS = wss_obs)
```

### Elbow Plot

```{r elbow-b, fig.height=5}
ggplot(elbow_B, aes(x = k, y = WSS)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 3) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow Plot – Observed-Only Distance Pipeline",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()

elbow_B$decrease <- c(NA, -diff(elbow_B$WSS))
elbow_B$pct_decrease <- c(NA, -diff(elbow_B$WSS) / head(elbow_B$WSS, -1) * 100)
knitr::kable(elbow_B, digits = 2, format = "markdown",
             caption = "WSS and marginal decrease per k (Observed-only pipeline)")
```

```{r select-k-B}
# ──────────────────────────────────────────────
# Adjust chosen_k_B based on YOUR elbow plot.
# ──────────────────────────────────────────────
chosen_k_B <- 4  # placeholder — update after inspecting the elbow plot
cat("Selected k for observed-only pipeline:", chosen_k_B, "\n")

km_B <- kmeans_results_B[[which(k_range == chosen_k_B)]]
```

### Cluster Visualisation in MDS Space

```{r mds-cluster-B, fig.height=7, fig.width=9}
scores_B <- as.data.frame(mds_coords[, 1:2])
colnames(scores_B) <- c("MDS1", "MDS2")
scores_B$cluster <- factor(km_B$cluster)

ggplot(scores_B, aes(x = MDS1, y = MDS2, colour = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(level = 0.68, linewidth = 0.8, linetype = "dashed") +
  labs(title = paste0("Observed-Only Pipeline Clusters (k = ", chosen_k_B,
                      ") in MDS Space"),
       colour = "Cluster") +
  theme_minimal()
```

```{r cluster-sizes-B}
cat("Cluster sizes (observed-only pipeline):\n")
print(table(km_B$cluster))
```

### Side-by-Side Elbow Comparison

```{r elbow-compare, fig.height=5}
elbow_combined <- rbind(
  data.frame(k = elbow_A$k, WSS = elbow_A$WSS, Pipeline = "Euclidean (imputed)"),
  data.frame(k = elbow_B$k, WSS = elbow_B$WSS, Pipeline = "Observed-only (MDS)")
)

# Normalise WSS to [0,1] per pipeline for visual comparison
elbow_combined <- elbow_combined %>%
  group_by(Pipeline) %>%
  mutate(WSS_norm = (WSS - min(WSS)) / (max(WSS) - min(WSS))) %>%
  ungroup()

ggplot(elbow_combined, aes(x = k, y = WSS_norm, colour = Pipeline)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Normalised Elbow Plots – Both Pipelines",
       x = "Number of Clusters (k)",
       y = "Normalised WSS (0 = best, 1 = worst)",
       colour = "Pipeline") +
  theme_minimal()
```

---

## Comparison of the Two Clusterings

### Proportion of Differing Assignments

Because $k$-means cluster labels are arbitrary (cluster "1" in Pipeline A need not
correspond to cluster "1" in Pipeline B), we first align labels using the
**best permutation** that maximises agreement. We then compute the proportion of
observations assigned differently.

```{r compare-clusters}
library(combinat)  # for permn()

labels_A <- km_A$cluster
labels_B <- km_B$cluster

# Use the maximum k from both pipelines for permutation alignment
k_max <- max(chosen_k_A, chosen_k_B)

# Best-permutation alignment
if (chosen_k_A == chosen_k_B) {
  perms <- combinat::permn(1:chosen_k_A)
  best_agree <- 0
  best_perm <- NULL
  for (perm in perms) {
    relabelled <- perm[labels_B]
    agreement <- sum(relabelled == labels_A)
    if (agreement > best_agree) {
      best_agree <- agreement
      best_perm <- perm
    }
  }
  labels_B_aligned <- best_perm[labels_B]
  prop_diff <- 1 - best_agree / length(labels_A)
} else {
  # If k differs, we simply report a cross-tabulation
  labels_B_aligned <- labels_B
  prop_diff <- NA
}

if (!is.na(prop_diff)) {
  cat(sprintf("Proportion of differing assignments: %.3f (%.1f%%)\n",
              prop_diff, prop_diff * 100))
} else {
  cat("Different k values chosen — direct proportion not meaningful.\n")
}
```

### Cross-Tabulation

```{r cross-tab}
if (chosen_k_A == chosen_k_B) {
  cross_tbl <- table(Euclidean = labels_A, ObservedOnly = labels_B_aligned)
} else {
  cross_tbl <- table(Euclidean = labels_A, ObservedOnly = labels_B)
}
knitr::kable(cross_tbl, format = "markdown",
             caption = "Cross-tabulation of cluster assignments (aligned labels)")
```

### Side-by-Side Visualisation

```{r side-by-side, fig.height=6, fig.width=12}
# Project Pipeline B assignments into Pipeline A's PCA space for direct comparison
compare_df <- data.frame(
  PC1 = scores_A$PC1,
  PC2 = scores_A$PC2,
  Euclidean = factor(labels_A),
  ObsOnly = if (chosen_k_A == chosen_k_B) factor(labels_B_aligned) else factor(labels_B)
)

p_A <- ggplot(compare_df, aes(x = PC1, y = PC2, colour = Euclidean)) +
  geom_point(alpha = 0.5, size = 1.5) +
  labs(title = paste0("Euclidean (k=", chosen_k_A, ")"), colour = "Cluster") +
  theme_minimal()

p_B <- ggplot(compare_df, aes(x = PC1, y = PC2, colour = ObsOnly)) +
  geom_point(alpha = 0.5, size = 1.5) +
  labs(title = paste0("Observed-Only (k=", chosen_k_B, ")"), colour = "Cluster") +
  theme_minimal()

gridExtra::grid.arrange(p_A, p_B, ncol = 2)
```

### Discussion

```{r discussion-placeholder, echo=FALSE, results='asis'}
cat("
**Similarity or difference?** The proportion of differing assignments (reported
above) quantifies the overall agreement. Values below ~15% suggest the two
pipelines recover essentially the same structure; values above ~30% indicate
meaningfully different clusterings.

**Possible reasons for differences:**

1. **Missing-data handling.** Column-mean imputation pulls missing values toward
   the centre of each variable's distribution, subtly shrinking within-cluster
   spread and potentially merging clusters that are close together. The
   renormalised distance avoids this by using only the coordinates that are
   actually observed, but at the cost of introducing variability in the effective
   dimensionality of each pairwise comparison.

2. **Geometry of the distance.** The Euclidean pipeline operates in the original
   $p$-dimensional feature space (after imputation). The observed-only pipeline
   operates in an MDS-embedded space of dimension $m < p$. MDS embedding is
   itself an approximation: it minimises a stress criterion and may distort local
   distances, particularly when the underlying distance matrix is non-Euclidean
   (as the renormalised distance can be).

3. **Scaling interaction.** Both pipelines standardise to unit variance, but
   the effective contribution of each variable differs. In Pipeline A, imputed
   values reduce a variable's apparent variance. In Pipeline B, variables with
   more missing values contribute to fewer pairwise distances, so their influence
   is naturally diluted without any artificial variance reduction.

4. **Sensitivity of $k$-means to initialisation.** Even with `nstart = 25`, the
   two pipelines operate in different spaces and may settle into different local
   optima. The elbow plots help verify that the chosen $k$ is robust rather than
   an artefact of a single initialisation.
")
```
