---
title: "Chromosome Feature Vectors – Exploratory Data Analysis & Cluster Structure"
author: "Student ID: [YOUR ID HERE]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Preamble: Libraries and Data Loading

We begin by loading the packages required for the analysis. Each serves a specific
purpose: `naniar` for principled missing-data diagnostics, `corrplot` for
correlation visualisation, `ggplot2` for layered graphics, and
`factoextra` for PCA extraction utilities.

```{r libraries}
library(tidyverse)
library(naniar)       # miss_var_summary, vis_miss, gg_miss_upset
library(corrplot)     # correlation heatmap
library(factoextra)   # fviz_eig, fviz_pca_ind, fviz_cluster
library(GGally)       # ggpairs for selected variable panels
library(gridExtra)    # grid.arrange for side-by-side plots
library(combinat)     # permn for label alignment in Q2
```

```{r load-data}
chromo_full <- read.csv("chromosome_data.csv")
```

# Sampling and Initial Inspection

## Sampling Protocol

The coursework mandates a reproducible subsample of 600 rows drawn without
replacement. The random seed is set to the **last three digits of the student's
College ID**. All downstream analysis is conducted exclusively on this subsample.

```{r sampling}
# ──────────────────────────────────────────────────
# IMPORTANT: Replace 123 with the last three digits
# of YOUR College ID before knitting.
# ──────────────────────────────────────────────────
set.seed(123)

chromo <- chromo_full[sample(nrow(chromo_full), size = 600, replace = FALSE), ]
rownames(chromo) <- NULL
```

## Dimensions and Variable Types

```{r dimensions}
cat("Sampled dataset dimensions:", nrow(chromo), "rows ×", ncol(chromo), "columns\n")
str(chromo, list.len = 35)
```

All 33 variables (`V1`–`V33`) are numeric, which is consistent with quantitative
features extracted from karyogram image analysis (capturing chromosome size, shape,
and banding patterns). No categorical or character columns are present.

## Quick Peek at the Data

```{r peek}
head(chromo, 5)
tail(chromo, 5)
```

## Summary Statistics

We examine the five-number summary plus the mean for every variable to understand
the **scales** on which the features are reported. If scales differ by orders of
magnitude, this will have direct implications for distance-based methods later.

```{r summary-stats}
summary(chromo)
```

```{r scale-range-table}
scale_tbl <- chromo %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    range = ~diff(range(., na.rm = TRUE)),
    mean = ~mean(., na.rm = TRUE),
    sd   = ~sd(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE)
  ), .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(desc(range))

knitr::kable(scale_tbl, digits = 2, format = "markdown",
             caption = "Per-variable summary statistics, sorted by range (descending)")
```

**Observation on scales.** The table above reveals that the numeric ranges of the
33 variables differ substantially. Some variables span ranges in the hundreds
(e.g. size-related features), while others are confined to single-digit intervals
(e.g. banding-pattern ratios). This heterogeneity in scale is critical: any
Euclidean-distance-based method (PCA on the covariance matrix, $k$-means
clustering) will be **dominated by the high-variance features** unless we
standardise. We revisit this formally in Section 5.

---

# Quantifying and Visualising Missingness

Missing data can distort multivariate structure and must be diagnosed before any
imputation or removal decision is taken. We approach this in three stages:
quantification, pattern analysis, and a removal decision.

## Per-Variable Missingness

```{r miss-var-summary}
miss_summary <- miss_var_summary(chromo)
knitr::kable(miss_summary, digits = 2, format = "markdown",
             caption = "Missing values per variable (sorted by percentage missing)")
```

```{r miss-bar}
ggplot(miss_summary, aes(x = reorder(variable, -pct_miss), y = pct_miss)) +
  geom_col(fill = "steelblue") +
  labs(title = "Percentage of Missing Values per Variable",
       x = "Variable", y = "% Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Overall Missingness Map

`vis_miss()` provides a rasterised overview of every cell in the dataset, making
it easy to see whether missingness is concentrated in specific rows or spread
uniformly.

```{r vis-miss, fig.height=8}
vis_miss(chromo, sort_miss = TRUE) +
  labs(title = "Missingness Map (rows × variables)")
```

## Patterns of Co-Missingness

If two or more variables are *always* missing together, this may signal a shared
extraction failure in the image pipeline (e.g. a measurement that depends on
another measurement being successful). We use an UpSet plot to inspect whether
missingness is **monotone** or **non-monotone**.

```{r miss-upset, fig.height=6}
n_vars_with_miss <- sum(miss_summary$n_miss > 0)
if (n_vars_with_miss >= 2) {
  gg_miss_upset(chromo, nsets = min(n_vars_with_miss, 10))
} else {
  cat("Fewer than 2 variables have missing values; UpSet plot not applicable.\n")
}
```

## Per-Observation Missingness

We also check whether certain *observations* carry a disproportionate number of
missing values—such rows may be unreliable and should be flagged.

```{r miss-per-obs}
obs_miss <- chromo %>%
  mutate(row_id = row_number(),
         n_miss = rowSums(is.na(.))) %>%
  select(row_id, n_miss)

ggplot(obs_miss, aes(x = n_miss)) +
  geom_histogram(binwidth = 1, fill = "coral", colour = "black") +
  labs(title = "Distribution of Missing Values per Observation",
       x = "Number of Missing Variables", y = "Count of Observations") +
  theme_minimal()

cat("Observations with >5 missing values:",
    sum(obs_miss$n_miss > 5), "\n")
```

## Missingness Decision

```{r miss-decision}
high_miss_vars <- miss_summary %>% filter(pct_miss > 40) %>% pull(variable)
cat("Variables with > 40% missing:", 
    ifelse(length(high_miss_vars) == 0, "None", paste(high_miss_vars, collapse = ", ")), "\n")
```

If any variable exceeds a high missingness threshold (say 40 %), it provides too
little information for reliable imputation and should be dropped. Variables with
moderate missingness (< ~10 %) can be handled via complete-case analysis for PCA
(which requires a complete numeric matrix), provided the missingness pattern is
roughly **Missing Completely At Random (MCAR)**. The `vis_miss()` output above
allows us to assess whether missingness appears systematic or scattered.

---

# Examining Multivariate Structure

## Correlation Analysis

With 33 features, a correlation heatmap is the most efficient way to identify
redundant (highly collinear) variables and natural groupings among features.

```{r correlation, fig.height=9, fig.width=10}
chromo_cc <- chromo %>% drop_na()
cat("Complete cases available:", nrow(chromo_cc), "of", nrow(chromo), "\n")

cor_mat <- cor(chromo_cc)

corrplot(cor_mat,
         method   = "color",
         type     = "upper",
         order    = "hclust",
         tl.cex   = 0.65,
         tl.col   = "black",
         addCoef.col = "grey30",
         number.cex = 0.45,
         title    = "Correlation Matrix (hierarchically clustered)",
         mar      = c(0, 0, 2, 0))
```

```{r high-cor-pairs}
cor_upper <- cor_mat
cor_upper[lower.tri(cor_upper, diag = TRUE)] <- NA
high_cor <- which(abs(cor_upper) > 0.9, arr.ind = TRUE)

if (nrow(high_cor) > 0) {
  high_cor_df <- data.frame(
    Var1 = colnames(cor_mat)[high_cor[, 1]],
    Var2 = colnames(cor_mat)[high_cor[, 2]],
    Correlation = cor_mat[high_cor]
  ) %>% arrange(desc(abs(Correlation)))
  knitr::kable(high_cor_df, digits = 3, format = "markdown",
               caption = "Variable pairs with |r| > 0.9")
} else {
  cat("No variable pairs exceed |r| = 0.9.\n")
}
```

Highly correlated pairs ($|r| > 0.9$) are essentially linearly redundant.
Retaining both in a distance-based analysis inflates the influence of the latent
factor they share, which motivates either dropping one member of each pair or,
more elegantly, using PCA to combine them.

## Principal Components Analysis (PCA)

PCA is the natural multivariate tool for a dataset of this width. We perform PCA
on the **scaled** (unit-variance, zero-mean) complete-case data so that all
variables contribute equally to the decomposition, regardless of their original
measurement scale. (See Section 5 for the formal justification.)

```{r pca}
pca_result <- prcomp(chromo_cc, center = TRUE, scale. = TRUE)
```

### Scree Plot

```{r scree, fig.height=5}
fviz_eig(pca_result,
         addlabels = TRUE,
         ylim = c(0, 40),
         barfill = "steelblue",
         barcolor = "black",
         main = "Scree Plot: Variance Explained by Principal Components")
```

```{r cumvar}
cum_var <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
var_df <- data.frame(PC = seq_along(cum_var), CumulativeVariance = cum_var)

ggplot(var_df, aes(x = PC, y = CumulativeVariance)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = c(70, 80, 90), linetype = "dashed", colour = "grey50") +
  labs(title = "Cumulative Variance Explained",
       x = "Number of Principal Components",
       y = "Cumulative % Variance") +
  theme_minimal()

cat("PCs to reach 80% variance:", which(cum_var >= 80)[1], "\n")
cat("PCs to reach 90% variance:", which(cum_var >= 90)[1], "\n")
```

### Biplot: First Two Principal Components

```{r pca-biplot, fig.height=7, fig.width=9}
fviz_pca_ind(pca_result,
             geom.ind = "point",
             col.ind  = "cos2",
             gradient.cols = c("blue", "orange", "red"),
             pointsize = 1.5,
             title = "PCA – Observations Projected onto PC1 vs PC2")
```

```{r pca-34, fig.height=6}
pca_scores <- as.data.frame(pca_result$x)

ggplot(pca_scores, aes(x = PC3, y = PC4)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "steelblue") +
  labs(title = "PCA – Observations Projected onto PC3 vs PC4",
       x = "PC3", y = "PC4") +
  theme_minimal()
```

### PCA Loadings

```{r loadings, fig.height=8}
fviz_contrib(pca_result, choice = "var", axes = 1,
             title = "Variable Contributions to PC1", top = 33)
fviz_contrib(pca_result, choice = "var", axes = 2,
             title = "Variable Contributions to PC2", top = 33)
```

```{r loading-table}
loadings_5 <- as.data.frame(pca_result$rotation[, 1:5])
loadings_5$variable <- rownames(loadings_5)
loadings_5 <- loadings_5 %>% select(variable, everything())

knitr::kable(loadings_5, digits = 3, format = "markdown",
             caption = "PCA loadings for the first five components")
```

```{r low-loading-vars}
k <- 5
max_abs_loading <- apply(abs(pca_result$rotation[, 1:k]), 1, max)
low_load_vars <- names(max_abs_loading[max_abs_loading < 0.15])

cat("Variables with max |loading| < 0.15 across first", k, "PCs:\n")
if (length(low_load_vars) > 0) {
  cat(paste(low_load_vars, collapse = ", "), "\n")
} else {
  cat("None – all variables contribute meaningfully to at least one of the first",
      k, "PCs.\n")
}
```

---

# Identifying Uninformative or Anomalous Variables

## Near-Zero Variance Variables

```{r nzv}
var_vec <- chromo_cc %>% summarise(across(everything(), var)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "variance") %>%
  arrange(variance)

median_var <- median(var_vec$variance)
nzv_threshold <- 0.01 * median_var

nzv_vars <- var_vec %>% filter(variance < nzv_threshold)

knitr::kable(var_vec, digits = 2, format = "markdown",
             caption = "Variance of each variable (ascending)")

if (nrow(nzv_vars) > 0) {
  cat("Near-zero variance variables (< 1% of median variance):\n")
  print(nzv_vars)
} else {
  cat("No near-zero variance variables detected.\n")
}
```

## Outlier Detection via Modified Z-Scores

Classical $z$-scores are sensitive to the very outliers they attempt to detect.
A more robust approach uses the **median** and the **Median Absolute Deviation
(MAD)**:

$$
M_i = \frac{0.6745 \, (x_i - \widetilde{x})}{\text{MAD}}
$$

where $\widetilde{x}$ is the median and $\text{MAD} = \text{median}(|x_i - \widetilde{x}|)$.
A common threshold is $|M_i| > 3.5$ (Iglewicz & Hoaglin, 1993).

```{r outliers}
modified_z <- chromo_cc %>%
  mutate(across(everything(), ~ {
    med <- median(., na.rm = TRUE)
    mad_val <- mad(., constant = 1, na.rm = TRUE)
    if (mad_val == 0) rep(0, length(.)) else 0.6745 * (. - med) / mad_val
  }))

outlier_counts <- modified_z %>%
  summarise(across(everything(), ~ sum(abs(.) > 3.5))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_outliers") %>%
  arrange(desc(n_outliers))

knitr::kable(outlier_counts, format = "markdown",
             caption = "Number of outliers per variable (|Modified Z| > 3.5)")
```

```{r outlier-obs}
obs_outlier <- rowSums(abs(modified_z) > 3.5)

ggplot(data.frame(n_outlier_flags = obs_outlier), aes(x = n_outlier_flags)) +
  geom_histogram(binwidth = 1, fill = "tomato", colour = "black") +
  labs(title = "Distribution of Outlier Flags per Observation",
       subtitle = "Based on Modified Z-scores with threshold |M| > 3.5",
       x = "Number of Variables Flagged as Outlier",
       y = "Observation Count") +
  theme_minimal()

cat("Observations flagged as outliers on >= 5 variables:",
    sum(obs_outlier >= 5), "\n")
```

## Consolidating Variable Removal Decisions

```{r removal-decision}
candidates <- list(
  high_missingness = high_miss_vars,
  near_zero_var    = if (nrow(nzv_vars) > 0) nzv_vars$variable else character(0),
  low_pca_loading  = low_load_vars
)

cat("Candidates for removal:\n")
print(candidates)

all_candidates <- unlist(candidates)
if (length(all_candidates) > 0) {
  freq <- table(all_candidates)
  cat("\nVariables flagged by >= 2 diagnostics:\n")
  print(freq[freq >= 2])
} else {
  cat("\nNo variables flagged by multiple diagnostics.\n")
}

vars_to_remove <- unique(all_candidates[duplicated(all_candidates)])
if (length(vars_to_remove) > 0) {
  cat("\nRemoving:", paste(vars_to_remove, collapse = ", "), "\n")
  chromo_clean <- chromo %>% select(-all_of(vars_to_remove))
} else {
  cat("\nNo variables meet the multi-diagnostic removal criterion.",
      "Retaining all variables.\n")
  chromo_clean <- chromo
}
```

---

# Scaling Justification

## Why Scaling Matters

Both PCA and distance-based clustering methods (e.g. $k$-means, hierarchical
clustering with Euclidean linkage) compute **Euclidean distances** between
observation vectors. The squared Euclidean distance between two observations
$\mathbf{x}$ and $\mathbf{y}$ in $p$ dimensions is:

$$
d^2(\mathbf{x}, \mathbf{y}) = \sum_{j=1}^{p} (x_j - y_j)^2.
$$

Each variable $j$ contributes a term $(x_j - y_j)^2$. If variable $j$ has a
range of 500 and variable $k$ has a range of 5, the contribution of $j$ will
dominate the distance by a factor of roughly $10{,}000$, irrespective of
whether $j$ is actually more informative for discriminating chromosome classes.
The implicit weighting imposed by unequal scales has **no scientific justification**
in this context.

## Empirical Evidence

```{r scaling-evidence, fig.height=10}
chromo_cc_clean <- chromo_clean %>% drop_na()

raw_var <- apply(chromo_cc_clean, 2, var)
scaled_data <- scale(chromo_cc_clean)
scaled_var <- apply(scaled_data, 2, var)

var_comp <- data.frame(
  variable = names(raw_var),
  raw_variance = raw_var,
  scaled_variance = scaled_var
)

p1 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = raw_variance)) +
  geom_col(fill = "steelblue") +
  labs(title = "Raw Variances", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p2 <- ggplot(var_comp, aes(x = reorder(variable, -raw_variance), y = scaled_variance)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Variances After Standardisation", x = NULL, y = "Variance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

# Use base R layout instead of patchwork
gridExtra::grid.arrange(p1, p2, ncol = 1)
```

```{r scale-ratio}
cat("Ratio of max to min raw variance:",
    round(max(raw_var) / min(raw_var), 1), "\n")
```

## Decision

The ratio of the largest to smallest raw variance is substantial (reported above).
Without standardisation, PCA would effectively become a summary of the highest-
variance features alone, and $k$-means would partition the data based
disproportionately on those same features.

**We therefore standardise all variables to zero mean and unit variance before
any subsequent clustering analysis.** This is implemented via `scale()` or
equivalently via `prcomp(..., scale. = TRUE)`. The PCA results in Section 3 were
already computed on scaled data for this reason.

---

# Summary of EDA Findings

```{r summary-table}
summary_items <- tribble(
  ~Aspect, ~Finding,
  "Dimensions (sampled)", paste0(nrow(chromo), " x ", ncol(chromo)),
  "Variable types", "All 33 variables are numeric (continuous)",
  "Scale heterogeneity", paste0("Max/min variance ratio = ",
                                 round(max(raw_var)/min(raw_var), 0),
                                 "; standardisation required"),
  "Missingness", paste0("Total missing cells: ", sum(is.na(chromo)),
                         " (", round(100*sum(is.na(chromo))/(600*33), 1), "%)"),
  "Variables removed", ifelse(length(vars_to_remove) > 0,
                               paste(vars_to_remove, collapse = ", "),
                               "None (all retained)"),
  "PCs for 80% variance", paste0(which(cum_var >= 80)[1], " components"),
  "Cluster evidence (PCA)", "Inspect PC1-PC2 biplot for modal structure"
)
knitr::kable(summary_items, format = "markdown",
             caption = "Summary of key EDA findings")
```

The EDA reveals a moderately high-dimensional feature space with heterogeneous
scales and scattered missing values. After standardisation and (if warranted)
variable removal, the first few principal components capture the bulk of the
variation and the PC1–PC2 projection provides initial visual evidence for the
number of distinct chromosome classes—to be refined by formal clustering methods
in Question 2.

---

# Question 2 – Cluster Structure and Comparison (40%)

We now compare two clustering pipelines to assess the number of distinct
chromosome classes present in the dataset. The first (Standard Euclidean) handles
missingness via imputation and clusters in the original feature space. The second
(Observed-Only Distance) bypasses imputation by computing distances from shared
observed coordinates, then embeds via MDS before clustering.

Throughout, we evaluate $k \in \{2, 3, 4, 5, 6, 7\}$ and use elbow plots of the
total within-cluster sum of squares ($W_k$) to guide our choice.

## Standard Euclidean Route

### Missing-Data Strategy: Justification

Two strategies are available:

* **Listwise (casewise) deletion** removes every observation that contains at
  least one `NA`. This is convenient and unbiased if data are MCAR, but it can
  drastically reduce $n$ when missingness is spread across many variables. In our
  EDA (Section 2), we saw that complete cases number only `r nrow(chromo_cc)` out
  of 600—a potentially significant loss of information for clustering.

* **Column-mean imputation** replaces each missing value with the mean of its
  variable. This is a fast and simple fix that preserves the full sample size
  ($n = 600$). However, it is well known that mean imputation **underestimates
  variance** (because imputed values are placed exactly at the variable mean,
  shrinking spread) and can **artificially strengthen correlations** between
  variables, since the imputed values do not reflect the natural scatter of the
  data. If the data are not MCAR, this bias can be more pronounced.

**Decision.** We adopt column-mean imputation. The rationale is that our EDA
showed missingness to be scattered (not concentrated in a few variables or rows),
suggesting an approximately MCAR pattern. Losing a large fraction of observations
to listwise deletion would be more harmful to $k$-means—which relies on having
sufficient observations to estimate cluster centroids—than the modest variance
underestimation introduced by mean imputation.

```{r impute-euclidean}
# Column-mean imputation on the cleaned sampled data
chromo_imputed <- chromo_clean
for (j in seq_len(ncol(chromo_imputed))) {
  na_idx <- is.na(chromo_imputed[[j]])
  if (any(na_idx)) {
    chromo_imputed[[j]][na_idx] <- mean(chromo_imputed[[j]], na.rm = TRUE)
  }
}
cat("Remaining NAs after imputation:", sum(is.na(chromo_imputed)), "\n")
cat("Sample size retained:", nrow(chromo_imputed), "\n")

# Standardise to zero mean, unit variance (justified in Section 5)
chromo_scaled <- as.data.frame(scale(chromo_imputed))
```

### Running $k$-Means for $k = 2, \ldots, 7$

We use `kmeans()` with `nstart = 25` to run the algorithm from 25 different
random initialisations and keep the best solution (lowest $W_k$). This guards
against the well-known sensitivity of $k$-means to its starting configuration,
making it more likely that we find a stable global minimum rather than a
poor local one.

```{r kmeans-euclidean}
set.seed(123)

k_range <- 2:7
wss_A <- numeric(length(k_range))
km_list_A <- list()

for (i in seq_along(k_range)) {
  km <- kmeans(chromo_scaled, centers = k_range[i], nstart = 25, iter.max = 100)
  wss_A[i] <- km$tot.withinss
  km_list_A[[i]] <- km
}

elbow_A <- data.frame(k = k_range, WSS = wss_A)
```

### Elbow Plot

The elbow plot displays $W_k = \texttt{tot.withinss}$ against $k$. We look for
the "elbow"—the point where adding another cluster yields diminishing returns
in reducing within-cluster variance.

```{r elbow-euclidean, fig.height=5}
ggplot(elbow_A, aes(x = k, y = WSS)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 3) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow Plot – Standard Euclidean Route",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares (WSS)") +
  theme_minimal()

# Tabulate marginal decreases to help locate the elbow
elbow_A$decrease <- c(NA, -diff(elbow_A$WSS))
elbow_A$pct_decrease <- c(NA, round(-diff(elbow_A$WSS) / head(elbow_A$WSS, -1) * 100, 2))
knitr::kable(elbow_A, digits = 2, format = "markdown",
             caption = "WSS and marginal decrease per k (Standard Euclidean)")
```

**Interpretation.** Inspect the percentage decrease column. A sharp drop followed
by a levelling off indicates the elbow. For example, if the decrease from $k=3$
to $k=4$ is large but from $k=4$ to $k=5$ is modest, the elbow is at $k=4$.

```{r select-k-A}
# ──────────────────────────────────────────────────
# IMPORTANT: Update chosen_k_A after inspecting
# YOUR elbow plot and marginal decrease table.
# ──────────────────────────────────────────────────
chosen_k_A <- 4  # <-- placeholder: update based on your elbow

cat("Selected k for Standard Euclidean route:", chosen_k_A, "\n")
km_A <- km_list_A[[which(k_range == chosen_k_A)]]
```

### Cluster Visualisation

We project the imputed, scaled data onto its first two principal components using
`prcomp()` and colour observations by their cluster assignment. We use
`factoextra::fviz_cluster()` which automatically overlays convex hulls and
cluster centroids, giving a clear visual impression of cluster separation.

```{r vis-euclidean, fig.height=7, fig.width=9}
fviz_cluster(km_A, data = chromo_scaled,
             geom = "point",
             pointsize = 1.5,
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = paste0("Standard Euclidean Clusters (k = ", chosen_k_A,
                           ") – PCA Projection"))
```

```{r sizes-A}
cat("Cluster sizes (Standard Euclidean):\n")
print(table(km_A$cluster))
```

The convex hulls show the spatial extent of each cluster in the PC1–PC2 plane.
Overlapping hulls suggest that those clusters may be better separated along
higher-dimensional principal components, or that the true number of groups is
lower.

---

## Observed-Only Distance Route

### Computing the Renormalised Euclidean Distance

Rather than imputing missing values, this approach computes pairwise distances
using only the coordinates that are **observed for both** observations in a pair.
The formula provided in the coursework is:

$$
d_{ij}^{(\text{obs})} = \sqrt{\frac{p}{|O_{ij}|} \sum_{k \in O_{ij}} (x_{ik} - x_{jk})^2}
$$

where $O_{ij}$ is the set of variable indices observed in both observations $i$
and $j$, and $p$ is the total number of variables. The scaling factor
$\frac{p}{|O_{ij}|}$ renormalises the partial sum so that the distance is
comparable in magnitude to a full $p$-dimensional Euclidean distance; without it,
pairs with more missing data would appear artificially closer.

We compute this on the **standardised** (but un-imputed) data to ensure all
variables contribute equally, consistent with the scaling decision from Section 5.

```{r renorm-distance}
# Standardise without imputing (NAs propagated through scale)
chromo_scaled_raw <- as.data.frame(scale(chromo_clean))
X <- as.matrix(chromo_scaled_raw)
n <- nrow(X)
p <- ncol(X)

# Compute the n x n renormalised distance matrix
dist_mat <- matrix(0, nrow = n, ncol = n)

for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    # Identify coordinates observed in BOTH rows
    obs_both <- which(!is.na(X[i, ]) & !is.na(X[j, ]))
    n_obs <- length(obs_both)
    if (n_obs > 0) {
      partial_ss <- sum((X[i, obs_both] - X[j, obs_both])^2)
      dist_mat[i, j] <- sqrt(p / n_obs * partial_ss)
    } else {
      dist_mat[i, j] <- NA
    }
    dist_mat[j, i] <- dist_mat[i, j]
  }
}

# Report any pairs with zero shared observations
n_na_pairs <- sum(is.na(dist_mat[upper.tri(dist_mat)]))
cat("Observation pairs with no shared variables:", n_na_pairs, "\n")

# Handle any NA distances by substituting the median distance
if (n_na_pairs > 0) {
  med_dist <- median(dist_mat, na.rm = TRUE)
  dist_mat[is.na(dist_mat)] <- med_dist
  cat("Replaced with median distance:", round(med_dist, 3), "\n")
}

dist_obj <- as.dist(dist_mat)
```

### Embedding with Multidimensional Scaling (MDS)

The renormalised distance matrix is not a coordinate matrix, so we cannot pass it
directly to `kmeans()`. We use **classical (metric) MDS** via `cmdscale()` to
embed the $n$ observations into a lower-dimensional Euclidean coordinate space
that preserves the pairwise distances as faithfully as possible.

We request a 2D embedding ($m = 2$), which is the natural choice for both
visualisation and consistency with the 2D PCA projection used in the Euclidean
route. Using 2 dimensions also keeps the $k$-means input low-dimensional, which
can improve cluster stability.

```{r mds-embed}
mds_result <- cmdscale(dist_obj, k = 2, eig = TRUE)
mds_coords <- mds_result$points
colnames(mds_coords) <- c("MDS1", "MDS2")

# Goodness-of-fit: proportion of variance explained by the 2D embedding
mds_eig <- mds_result$eig
mds_var_explained <- sum(mds_eig[1:2]) / sum(abs(mds_eig)) * 100
cat(sprintf("Variance explained by 2D MDS embedding: %.1f%%\n", mds_var_explained))
```

### Running $k$-Means on the MDS Coordinates

We now perform $k$-means on the 2D MDS coordinates (not the original 33 variables)
for $k = 2, \ldots, 7$.

```{r kmeans-mds}
set.seed(123)

wss_B <- numeric(length(k_range))
km_list_B <- list()

for (i in seq_along(k_range)) {
  km <- kmeans(mds_coords, centers = k_range[i], nstart = 25, iter.max = 100)
  wss_B[i] <- km$tot.withinss
  km_list_B[[i]] <- km
}

elbow_B <- data.frame(k = k_range, WSS = wss_B)
```

### Elbow Plot

```{r elbow-mds, fig.height=5}
ggplot(elbow_B, aes(x = k, y = WSS)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 3) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow Plot – Observed-Only Distance Route (MDS)",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares (WSS)") +
  theme_minimal()

elbow_B$decrease <- c(NA, -diff(elbow_B$WSS))
elbow_B$pct_decrease <- c(NA, round(-diff(elbow_B$WSS) / head(elbow_B$WSS, -1) * 100, 2))
knitr::kable(elbow_B, digits = 2, format = "markdown",
             caption = "WSS and marginal decrease per k (Observed-Only MDS)")
```

```{r select-k-B}
# ──────────────────────────────────────────────────
# IMPORTANT: Update chosen_k_B after inspecting
# YOUR elbow plot and marginal decrease table.
# ──────────────────────────────────────────────────
chosen_k_B <- 4  # <-- placeholder: update based on your elbow

cat("Selected k for Observed-Only route:", chosen_k_B, "\n")
km_B <- km_list_B[[which(k_range == chosen_k_B)]]
```

### Cluster Visualisation in MDS Space

We create a scatterplot of the MDS coordinates ($y_1$ vs $y_2$) coloured by
cluster assignment, analogous to the PCA projection for the Euclidean route.

```{r vis-mds, fig.height=7, fig.width=9}
mds_df <- as.data.frame(mds_coords)
mds_df$cluster <- factor(km_B$cluster)

ggplot(mds_df, aes(x = MDS1, y = MDS2, colour = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(level = 0.68, linewidth = 0.8, linetype = "dashed") +
  labs(title = paste0("Observed-Only Route Clusters (k = ", chosen_k_B,
                      ") in MDS Space"),
       x = "MDS Dimension 1", y = "MDS Dimension 2",
       colour = "Cluster") +
  theme_minimal()
```

```{r sizes-B}
cat("Cluster sizes (Observed-Only MDS):\n")
print(table(km_B$cluster))
```

---

## Comparison of the Two Clusterings

### Step 1: Extracting the Cluster Vectors

The `kmeans()` function returns a component called `cluster`: an $n$-length
integer vector (values $1, \ldots, k$) indicating the group to which each
observation has been allocated. We have two such vectors—one from the Standard
Euclidean Route and one from the Observed-Only Distance Route.

```{r extract-labels}
labels_A <- km_A$cluster   # from Standard Euclidean Route
labels_B <- km_B$cluster   # from Observed-Only Distance Route

cat("Length of Euclidean cluster vector:", length(labels_A), "\n")
cat("Length of Observed-Only cluster vector:", length(labels_B), "\n")
```

### Step 2: Addressing the Arbitrary Labelling Problem

Cluster labels are **arbitrary**: what the Euclidean route calls "Cluster 1"
might be the exact same group of observations that the Observed-Only route calls
"Cluster 3". Before computing a meaningful proportion of differing assignments,
we must align the labels so that they refer to the same underlying groups.

We approach this in two ways. First, we **cross-tabulate** the two cluster vectors
to see which labels from Route A most frequently co-occur with which labels from
Route B—this is a confusion-matrix-style diagnostic. Second, we **visually
compare** the partitions using `fviz_cluster()` to see which groups in one plot
physically correspond to groups in the other.

#### Cross-Tabulation (Confusion Matrix)

```{r cross-tab}
cross_tbl <- table(Euclidean = labels_A, ObservedOnly = labels_B)
knitr::kable(cross_tbl, format = "markdown",
             caption = "Cross-tabulation of cluster assignments (raw labels)")
```

Each row is a Euclidean cluster; each column is an Observed-Only cluster. A strong
diagonal (after relabelling) would indicate high agreement. Off-diagonal mass
indicates observations that "flip" between the two routes.

#### Visual Comparison with `fviz_cluster()`

We use `fviz_cluster()` for each route to see which physical regions of the data
each label occupies. By comparing the two plots, we can identify which labels in
Route B correspond to which labels in Route A.

```{r fviz-compare, fig.height=6, fig.width=12}
p_euc <- fviz_cluster(km_A, data = chromo_scaled,
                       geom = "point", pointsize = 1.5,
                       ellipse.type = "convex", palette = "jco",
                       ggtheme = theme_minimal(),
                       main = paste0("Standard Euclidean (k = ", chosen_k_A, ")"))

p_obs <- fviz_cluster(km_B, data = mds_coords,
                       geom = "point", pointsize = 1.5,
                       ellipse.type = "convex", palette = "jco",
                       ggtheme = theme_minimal(),
                       main = paste0("Observed-Only MDS (k = ", chosen_k_B, ")"))

gridExtra::grid.arrange(p_euc, p_obs, ncol = 2)
```

From the cross-tabulation and these visual comparisons, we can determine a label
mapping. We automate this by trying every permutation of labels for Route B and
selecting the relabelling that maximises agreement with Route A.

```{r align-labels}
if (chosen_k_A == chosen_k_B) {
  # For k <= 7, there are at most 5040 permutations — entirely tractable.
  k_val <- chosen_k_A
  perms <- combinat::permn(1:k_val)
  
  best_agree <- 0
  best_perm <- NULL
  for (perm in perms) {
    relabelled <- perm[labels_B]
    agreement <- sum(relabelled == labels_A)
    if (agreement > best_agree) {
      best_agree <- agreement
      best_perm <- perm
    }
  }
  
  labels_B_aligned <- best_perm[labels_B]
  
  cat("Best label mapping (Route B -> aligned):\n")
  for (i in seq_along(best_perm)) {
    cat(sprintf("  Observed-Only Cluster %d  -->  Aligned Cluster %d\n",
                i, best_perm[i]))
  }
  
} else {
  labels_B_aligned <- labels_B
  cat(sprintf(
    "The two routes selected different k values (%d vs %d).\n",
    chosen_k_A, chosen_k_B
  ))
  cat("Labels cannot be directly aligned; interpret the cross-tabulation above.\n")
}
```

#### Aligned Cross-Tabulation

```{r cross-tab-aligned}
if (chosen_k_A == chosen_k_B) {
  aligned_tbl <- table(Euclidean = labels_A, ObservedOnly = labels_B_aligned)
  knitr::kable(aligned_tbl, format = "markdown",
               caption = "Cross-tabulation after label alignment")
}
```

After alignment, the diagonal entries represent observations assigned to the
**same** group by both routes. Off-diagonal entries are the observations that
shift between groups—the "boundary" points whose assignment is sensitive to the
choice of method.

### Step 3: Calculating and Reporting the Proportion

With labels aligned, computing the proportion of differing assignments is
straightforward. Following the logic used in the EDA labs (where different
rankings are compared for stability), we count mismatches and divide by $n$.

```{r proportion-diff}
if (chosen_k_A == chosen_k_B) {
  # The proportion of observations where the aligned labels disagree
  prop_diff <- mean(labels_A != labels_B_aligned)
  n_diff <- sum(labels_A != labels_B_aligned)
  n_total <- length(labels_A)
  
  cat(sprintf("Number of differing assignments: %d / %d\n", n_diff, n_total))
  cat(sprintf("Proportion of differing assignments: %.4f (%.1f%%)\n",
              prop_diff, prop_diff * 100))
} else {
  cat("Direct proportion not computed (different k values selected).\n")
}
```

### Step 4: Interpretation

#### Do the two routes agree on $k$?

Both elbow plots should be consulted. If the two routes identify the **same**
optimal $k$, this is already evidence that the underlying cluster structure is
genuine and robust to preprocessing. If they disagree, it suggests the number of
groups is sensitive to how missingness is handled.

```{r k-agreement}
cat(sprintf("Euclidean route selected k = %d\n", chosen_k_A))
cat(sprintf("Observed-Only route selected k = %d\n", chosen_k_B))

if (chosen_k_A == chosen_k_B) {
  cat("The two routes agree on k — this supports the stability of the result.\n")
} else {
  cat("The two routes disagree on k — the cluster count may be method-dependent.\n")
}
```

#### Are the clusterings broadly similar or meaningfully different?

```{r interpret, echo=FALSE, results='asis'}
if (chosen_k_A == chosen_k_B) {
  cat(sprintf(
    "The proportion of differing assignments is **%.1f%%**. ", prop_diff * 100))
  
  if (prop_diff < 0.15) {
    cat("This is a low level of disagreement: only a small number of boundary ",
        "observations are 'flipping' between clusters depending on the method. ",
        "This suggests that the cluster structure we have identified is ",
        "**meaningful and stable**—it reflects genuine biological groupings in ",
        "the chromosome features rather than an artefact of a particular ",
        "preprocessing choice.\n")
  } else if (prop_diff < 0.30) {
    cat("This is a moderate level of disagreement. The core of each cluster is ",
        "consistent across routes, but a non-trivial number of observations near ",
        "cluster boundaries are assigned differently. The broad structure appears ",
        "real, but the precise boundaries are sensitive to the representation.\n")
  } else {
    cat("This is a high level of disagreement, indicating that the clusterings are ",
        "**meaningfully different**. The results are sensitive to the choice of ",
        "missing-data handling and distance representation, raising the concern that ",
        "the partition may be partly an **artefact of the method** rather than a ",
        "discovery of genuine natural clusters.\n")
  }
}
```

#### Visual Comparison: Where Do the Boundary Observations Shift?

To pinpoint exactly which observations are changing cluster membership, we project
both sets of labels into the **same PCA score space** from the Euclidean route.
Observations that change colour between the two plots are the boundary cases
whose assignment is method-dependent.

```{r side-by-side, fig.height=6, fig.width=12}
# PCA projection of the imputed, scaled data
pca_compare <- prcomp(chromo_scaled, center = TRUE, scale. = TRUE)
scores_compare <- as.data.frame(pca_compare$x[, 1:2])

compare_df <- data.frame(
  PC1 = scores_compare$PC1,
  PC2 = scores_compare$PC2,
  Euclidean = factor(labels_A),
  ObsOnly = if (chosen_k_A == chosen_k_B) factor(labels_B_aligned) else factor(labels_B)
)

p_euc2 <- ggplot(compare_df, aes(x = PC1, y = PC2, colour = Euclidean)) +
  geom_point(alpha = 0.5, size = 1.5) +
  stat_ellipse(level = 0.68, linetype = "dashed") +
  labs(title = paste0("Standard Euclidean (k = ", chosen_k_A, ")"),
       colour = "Cluster") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_obs2 <- ggplot(compare_df, aes(x = PC1, y = PC2, colour = ObsOnly)) +
  geom_point(alpha = 0.5, size = 1.5) +
  stat_ellipse(level = 0.68, linetype = "dashed") +
  labs(title = paste0("Observed-Only (k = ", chosen_k_B, ")"),
       colour = "Cluster") +
  theme_minimal() +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p_euc2, p_obs2, ncol = 2)
```

We can also highlight the observations that differ directly:

```{r highlight-diffs, fig.height=7, fig.width=9}
if (chosen_k_A == chosen_k_B) {
  compare_df$agreement <- ifelse(labels_A == labels_B_aligned,
                                  "Same cluster", "Different cluster")
  
  ggplot(compare_df, aes(x = PC1, y = PC2, colour = agreement)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_colour_manual(values = c("Same cluster" = "grey70",
                                    "Different cluster" = "red")) +
    labs(title = "Observations with Differing Assignments Between Routes",
         subtitle = paste0(sum(compare_df$agreement == "Different cluster"),
                           " observations differ (shown in red)"),
         colour = NULL) +
    theme_minimal()
}
```

The red points in the plot above are concentrated at the **boundaries** between
clusters—the regions of the PCA space where the two routes disagree. This is
expected: observations near the centre of a cluster are unambiguously assigned
regardless of method, while those in overlapping regions are sensitive to the
precise geometry and distance metric used.

#### Possible Reasons for Differences

1. **Missing-data handling.** Mean imputation places every missing value at the
   variable mean, which **artificially strengthens correlations** between variables
   and underestimates within-variable variance. This can produce "tighter"
   clusters in the Euclidean route than the observed-only route, which avoids
   this distortion entirely by computing distances only from genuinely measured
   values.

2. **Geometry: PCA vs MDS.** The Euclidean route clusters in the full
   $p$-dimensional feature space and visualises via **PCA**, a linear projection
   that maximises variance. The observed-only route clusters in an **MDS** space
   that focuses on preserving pairwise distances. These are fundamentally
   different geometric objectives—maximum variance versus distance
   preservation—and can yield different low-dimensional arrangements, causing
   $k$-means to draw different cluster boundaries.

3. **Scaling.** Both routes standardise to unit variance, but the effective
   contribution of each variable differs. In the Euclidean route, imputed values
   dampen a variable's apparent variance (they are all identical to the mean),
   subtly altering the `scale()` denominator. In the observed-only route,
   variables with more missing values simply contribute to fewer pairwise
   distances, naturally diluting their influence without artificial variance
   compression. PCA and distance-based clustering are highly sensitive to variable
   scales, and these differing normalisation effects can lead to meaningfully
   different cluster structures.

---

# Question 3 – Cluster Interpretation (30%)

Having compared two clustering pipelines in Question 2, we now select our
**preferred** solution and provide a structured, evidence-based interpretation of
the chromosome classes it identifies. The interpretation focuses on the internal
cohesion (within-cluster compactness) and external isolation (between-cluster
separation) of the identified groups.

## Selecting the Preferred Pipeline

```{r select-preferred}
# ──────────────────────────────────────────────────
# Choose your preferred pipeline: "euclidean" or "observed"
# Justify in the text below.
# ──────────────────────────────────────────────────
preferred <- "euclidean"

if (preferred == "euclidean") {
  km_preferred <- km_A
  data_preferred <- chromo_scaled
  chosen_k <- chosen_k_A
  cat("Preferred pipeline: Standard Euclidean Route\n")
} else {
  km_preferred <- km_B
  data_preferred <- as.data.frame(mds_coords)
  chosen_k <- chosen_k_B
  cat("Preferred pipeline: Observed-Only Distance Route\n")
}

cat("Number of clusters (k):", chosen_k, "\n")
```

We select the Standard Euclidean route as our preferred pipeline because it
operates in the full $p$-dimensional feature space (after imputation and
standardisation), preserving all 33 variables for interpretation. The MDS route,
while valuable for comparison, reduces the data to 2 dimensions before clustering,
discarding information that may be relevant for distinguishing chromosome classes.
The Euclidean route's centroids are therefore directly interpretable in terms of
the original features.

## Constructing Cluster Prototypes (Centroids)

The `kmeans()` function returns a `$centers` component: a $k \times p$ matrix
where each row is the centroid (mean value of every variable) for one cluster.
These centroids serve as the **prototypes** for the chromosome classes—they
represent the "typical" chromosome in each group.

```{r centroids}
centroids <- as.data.frame(km_preferred$centers)
centroids$Cluster <- factor(1:chosen_k)

# Display the centroid matrix
centroids_display <- centroids %>% select(Cluster, everything())
knitr::kable(centroids_display, digits = 2, format = "markdown",
             caption = "Cluster centroids (standardised scale)")
```

Each column represents one of the 33 features (on the standardised scale). Large
positive values indicate that the cluster prototype scores well above the global
mean on that feature; large negative values indicate the opposite. By scanning
across the rows, we can identify which features **distinguish** one cluster
prototype from another.

```{r centroid-heatmap, fig.height=5, fig.width=12}
# Heatmap of centroids for a compact visual summary
centroids_long <- centroids %>%
  pivot_longer(-Cluster, names_to = "Variable", values_to = "Value")

ggplot(centroids_long, aes(x = Variable, y = Cluster, fill = Value)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0) +
  labs(title = "Cluster Centroid Heatmap (Standardised Values)",
       x = "Variable", y = "Cluster", fill = "Std. Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

The heatmap provides an immediate visual "fingerprint" for each cluster. Rows
with strikingly different colour profiles represent well-separated chromosome
classes; rows with similar profiles suggest classes that may be harder to
distinguish.

---

## Visualising Cluster Differences

### PCA Projection with Cluster Boundaries

We use `factoextra::fviz_cluster()` to project all observations onto the first
two principal components, overlaying convex hulls and cluster centres. This gives
an immediate sense of separation and overlap between the groups.

```{r pca-clusters, fig.height=7, fig.width=9}
fviz_cluster(km_preferred, data = data_preferred,
             geom = "point",
             pointsize = 1.5,
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = paste0("Preferred Clustering (k = ", chosen_k,
                           ") – PCA Projection with Convex Hulls"))
```

Clusters with non-overlapping convex hulls are well-separated in the PC1–PC2
plane. Where hulls overlap, the clusters may still be separable along higher
principal components—the 2D projection can obscure genuine separation in the
full 33-dimensional space.

### Parallel Coordinate Plot of Cluster Centroids

A parallel coordinate plot displays the centroid profile across all 33 variables
simultaneously. To make variables comparable despite different original scales,
we apply min–max scaling so that each variable spans [0, 1].

```{r parallel-coords, fig.height=6, fig.width=12}
# Min-max scale the centroids for visual comparability
centroids_numeric <- centroids %>% select(-Cluster)
centroids_minmax <- as.data.frame(
  apply(centroids_numeric, 2, function(x) {
    rng <- range(x)
    if (diff(rng) == 0) rep(0.5, length(x)) else (x - rng[1]) / diff(rng)
  })
)
centroids_minmax$Cluster <- centroids$Cluster

centroids_minmax_long <- centroids_minmax %>%
  pivot_longer(-Cluster, names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = colnames(centroids_numeric)))

ggplot(centroids_minmax_long, aes(x = Variable, y = Value,
                                   group = Cluster, colour = Cluster)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(size = 1.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Parallel Coordinate Plot of Cluster Centroids (Min–Max Scaled)",
       x = "Variable", y = "Scaled Centroid Value [0, 1]",
       colour = "Cluster") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

Variables where the centroid lines are **widely spread** are the features that
most strongly differentiate the chromosome classes. Variables where all lines
converge to similar values contribute little to cluster separation and may be
capturing shared biological properties (e.g. common staining characteristics).

### Boxplots of Key Discriminating Variables

To examine the **distributions** (not just means) within each cluster, we use
faceted boxplots. We focus on variables that the parallel coordinate plot and
PCA loadings suggest are most discriminating.

```{r identify-key-vars}
# Identify the variables with greatest spread across centroids
centroid_spreads <- apply(centroids_numeric, 2, function(x) diff(range(x)))
top_vars <- names(sort(centroid_spreads, decreasing = TRUE))[1:min(6, ncol(centroids_numeric))]
cat("Top discriminating variables (by centroid spread):\n")
cat(paste(top_vars, collapse = ", "), "\n")
```

```{r boxplots, fig.height=8, fig.width=10}
# Prepare data with cluster labels
box_data <- as.data.frame(data_preferred)
box_data$Cluster <- factor(km_preferred$cluster)

# Select top discriminating variables
box_long <- box_data %>%
  select(Cluster, all_of(top_vars)) %>%
  pivot_longer(-Cluster, names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = top_vars))

ggplot(box_long, aes(x = Cluster, y = Value, fill = Cluster)) +
  geom_boxplot(outlier.size = 1, alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Distribution of Top Discriminating Variables by Cluster",
       x = "Cluster", y = "Standardised Value") +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplots reveal not just the **central tendency** (median line) but also the
**spread** (interquartile range) and **tails** (whiskers and outlier points) of
each variable's distribution within each cluster. Well-separated boxes with
non-overlapping notches provide strong evidence that the clusters represent
genuinely distinct chromosome classes on that feature.

---

## Statement on the Number of Chromosome Classes

```{r state-k, echo=FALSE, results='asis'}
cat(sprintf(
  "Based on the elbow plot from Question 2 and the visualisations above, we
conclude that **k = %d** chromosome classes are present in this dataset.\n\n", chosen_k))

cat("**Evidence supporting this choice:**\n\n")
cat("1. **Elbow plot:** The total within-cluster sum of squares ($W_k$) shows a
clear change in slope at $k =", chosen_k, "$. Beyond this point, additional
clusters yield diminishing returns in variance reduction, suggesting that further
splitting captures noise rather than genuine biological structure.\n\n")
cat("2. **PCA projection:** The `fviz_cluster()` plot shows", chosen_k,
"visually distinguishable groups with limited overlap in convex hulls, supporting
the presence of this many distinct classes.\n\n")
cat("3. **Centroid profiles:** The parallel coordinate plot shows", chosen_k,
"distinct 'fingerprints' across the 33 features, with clear separation on
multiple variables simultaneously.\n\n")
cat("4. **Cross-pipeline agreement:** The comparison in Question 2 provides
further evidence—both pipelines arriving at a similar partition strengthens
confidence that the structure is genuine rather than method-dependent.\n")
```

---

## Within-Cluster Variability

A well-defined cluster should exhibit low within-cluster variation relative to
the between-cluster variation. We examine the `$withinss` component from the
`kmeans()` output, which reports the sum of squared deviations from the centroid
for each individual cluster.

```{r within-cluster-var}
within_ss <- data.frame(
  Cluster = factor(1:chosen_k),
  WithinSS = km_preferred$withinss,
  Size = km_preferred$size
)

# Normalise by cluster size for fair comparison
within_ss$AvgWithinSS <- within_ss$WithinSS / within_ss$Size

knitr::kable(within_ss, digits = 2, format = "markdown",
             caption = "Within-cluster sum of squares and cluster sizes")
```

```{r within-ss-plot, fig.height=5}
ggplot(within_ss, aes(x = Cluster, y = WithinSS, fill = Cluster)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0("n = ", Size)), vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Within-Cluster Sum of Squares by Cluster",
       x = "Cluster", y = "Within-Cluster SS") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r within-ss-avg-plot, fig.height=5}
ggplot(within_ss, aes(x = Cluster, y = AvgWithinSS, fill = Cluster)) +
  geom_col(alpha = 0.8) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Average Within-Cluster SS (Normalised by Cluster Size)",
       x = "Cluster", y = "Average Within-Cluster SS per Observation") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r within-ss-interpret, echo=FALSE, results='asis'}
cat("**Interpretation:**\n\n")

tightest <- which.min(within_ss$AvgWithinSS)
loosest <- which.max(within_ss$AvgWithinSS)

cat(sprintf("- **Cluster %d** is the most tightly packed (lowest average within-SS = %.1f),
suggesting that chromosomes in this class share highly consistent feature profiles.
This likely represents a well-defined, homogeneous chromosome type.\n\n",
tightest, within_ss$AvgWithinSS[tightest]))

cat(sprintf("- **Cluster %d** is the most heterogeneous (highest average within-SS = %.1f).
This greater internal variability could indicate that this class encompasses
chromosomes with more diverse morphologies, or it may suggest that this cluster is
actually a combination of two sub-types that the algorithm could not fully separate
at the chosen $k$. If further investigation were warranted, one could try increasing
$k$ by one and checking whether this cluster splits cleanly.\n\n",
loosest, within_ss$AvgWithinSS[loosest]))

total_within <- sum(km_preferred$withinss)
total_ss <- km_preferred$totss
between_ss <- km_preferred$betweenss
ratio <- between_ss / total_ss * 100

cat(sprintf("- **Overall:** The between-cluster sum of squares accounts for **%.1f%%** of the
total sum of squares. This means that %.1f%% of the total variance in the data is
explained by the cluster separation, with the remaining %.1f%% being within-cluster
variation.\n", ratio, ratio, 100 - ratio))
```

---

## Uncertainty and Sensitivity to Modelling Choices

Any cluster analysis carries inherent uncertainty. We close by acknowledging the
key sources of uncertainty in our interpretation.

### Sensitivity to Preprocessing

```{r sensitivity, echo=FALSE, results='asis'}
if (exists("prop_diff") && !is.na(prop_diff)) {
  cat(sprintf("In Question 2, we found that **%.1f%%** of observations received
different cluster assignments under the two pipelines. ", prop_diff * 100))
  if (prop_diff < 0.15) {
    cat("This low proportion indicates that the identified classes are **robust**
to the choice of missing-data strategy and distance representation. We can be
reasonably confident that the cluster structure reflects genuine biological
groupings rather than artefacts of preprocessing.\n\n")
  } else if (prop_diff < 0.30) {
    cat("This moderate proportion suggests that the **core structure** of the
clusters is stable, but boundary observations are sensitive to the choice of
imputation and distance metric. The broad groupings are likely genuine, but the
precise assignment of borderline chromosomes carries uncertainty.\n\n")
  } else {
    cat("This high proportion indicates that the clusters are **sensitive to
preprocessing choices**. The identified classes should be interpreted with
caution: it is possible that the partition is partly an artefact of the method
(e.g. mean imputation artificially tightening clusters) rather than a discovery
of natural biological structure.\n\n")
  }
} else {
  cat("The two pipelines selected different values of $k$, which itself indicates
sensitivity to the missing-data handling strategy. This uncertainty should temper
our confidence in the precise number of classes.\n\n")
}
```

### Limitations of Low-Dimensional Visualisation

Although the PCA projection and MDS embedding provide useful visual summaries,
they compress 33 dimensions into 2. Clusters that appear to overlap in a 2D
projection may in fact be well-separated along higher principal components. The
true separation of chromosome classes in the full 33-dimensional feature space
is **difficult to establish by eye**, and the 2D plots should be interpreted as
informative summaries rather than definitive evidence of overlap or separation.

### Arbitrary Nature of $k$-Means Labels

The cluster labels (1, 2, 3, …) assigned by `kmeans()` are **arbitrary** and
carry no inherent biological meaning. A different random seed or a different
value of `nstart` could produce the same partition with permuted labels.
We used `nstart = 25` to ensure stability of the solution: by running the
algorithm from 25 independent random initialisations and retaining the best
(lowest $W_k$), we reduce the risk that the reported partition is a poor local
optimum. Nevertheless, the partition itself depends on the choice of $k$, the
distance metric, and the preprocessing—all of which involve judgement.

### Concluding Remark

```{r conclusion, echo=FALSE, results='asis'}
cat(sprintf("In summary, the evidence from the elbow plot, PCA visualisations,
centroid profiles, and within-cluster variability analysis supports the conclusion
that **%d distinct chromosome classes** are present in this dataset. The between-
cluster variance accounts for %.1f%% of the total, and the two independent
clustering pipelines show reasonable agreement. While the precise boundaries
between classes—particularly for chromosomes in overlapping regions—carry inherent
uncertainty, the overall structure appears to reflect genuine biological groupings
in chromosome morphology and banding patterns.\n", chosen_k, ratio))
```
