---
title: \normalsize Assessment 4 \hfill MATH70094 Programming for Data Science \hfill Autumn 2025
    \rule{\linewidth}{0.5mm}
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Graham Dib and CID: 01709270

# Question 1 (75 marks)

**The only non-base-R packages you are allowed to load in this question
are `igraph`, `R6` and `testthat`.**

Your task is to build an R package that implements and analyses
discrete-time Markov chains using R6 classes.

## Background

Let $(X_t)_{t\in\mathbb{N}}$ be a *discrete-time time-homogeneous Markov
chain* with finite state space $S=\{1,\dots,K\}$ for $K\in\mathbb{N}$.
Let $P\in\mathbb{R}^{K\times K}$ be the transition matrix of the Markov
chain which contains the transition probabilities
$P_{ij}=Pr(X_{t+1}=j|X_t=i)$. We mention here the following facts from
the theory of Markov chains:

(1) If $\pi\in\mathbb{R}^K$ is a vector describing the initial
    distribution of the chain over the state space $S$ (so the
    components of $\pi$ are between 0 and 1 and sum to 1), then the
    marginal distribution $\pi_t$ of $X_t$ over $S$ for $t\in\mathbb{N}$
    can be obtained from $\pi_t = \pi P^t$, where $P^t=P P \cdots P$ is
    the $t$-fold matrix product of $P$.
(2) A distribution $\pi$ is called *stationary* if $\pi=\pi P$.
(3) If there exists a unique stationary distribution $\pi$ (i.e. if the
    chain is irreducible since we are on a finite state space), then the
    *power method* says that for any initial distribution $\pi_0$ we
    have convergence of $\pi_0 P^t$ to $\pi$ as $t$ tends to infinity.
(4) The *ergodic theorem* for Markov chains says that if there exists a
    unique stationary distribution $\pi$, then for any initial state
    $X_0=i\in\{1,\dots,K\}$ we have
    $N^{-1}\sum_{t=1}^N \mathbf{1}(X_t=i) \rightarrow \pi(i)$ as $N$
    tends to infinity where $\mathbf{1}(X_t=i)$ equals 1 when $X_t=i$
    and zero otherwise, and where $\pi(i)$ is the $\pi$-probability of
    being in state $i$. This means that the relative frequency of being
    in state $i$ converges along any path to $\pi(i)$.

## Code clarity (5 marks)

There is a famous saying among software developers that code is read
more often than it is written. Marks will be awarded (or not awarded)
based on the clarity of the code and appropriate use of comments.

## Part A (18 marks)

Create a file `markovChain.R` containing an R6 class `MarkovChain` with
at least the following fields and methods:

(1) fields:

-   `initial_state` - an index from $1,\dots,K$ with the initial state
-   `P` - a quadratic transition matrix of dimension $K\times K$
    (entries must be positive, between 0 and 1 and the row sums should
    be 1)

(2) methods:

-   `initialize` - takes as inputs a transition matrix `P` and an
    `initial_state` and sets the fields in (1)
-   `simulate` - takes as inputs a number of steps `n_steps` and a
    number of distinct simulation paths `n_paths`, simulates `n_path`
    independent trajectories of length `n_steps` all starting from
    `initial_state`, returns a convenient R object with the simulation
    results
-   `marginal_distribution` - takes as input a time `t`, returns the
    marginal distribution at time `t` when starting from the initial
    distribution with point mass at `initial_state`
-   `empirical_marginal_distribution` - takes as inputs paths of a
    Markov chain simulation (obtained from `simulate`) and returns the
    empirical distribution at each of the time points (i.e. for each
    state at this time point the relative frequency along the paths)
-   `stationary_distribution` - takes as inputs a tolerance value `tol`
    and a maximum number of iterations `max_iter` and returns an
    estimate of the stationary distribution (assuming there exists one)
    by the power method starting from the initial distribution with
    point mass at `initial_state`
-   `visualise` - takes no inputs and prints the directed weighted graph
    corresponding to the transition matrix using
    `igraph::graph_from_adjacency_matrix()` from the `igraph` package
    and plots it; edge labels reflect transition probabilities.

**Observe the following:**

-   Include appropriate (but brief) documentation for all methods.
-   Use defensive programming techniques to check for incorrect input.
-   Fields and methods may be public or private.
-   You may add additional fields and methods if you think this is
    helpful.
-   Part B gives marks for appropriate unit tests. Try to follow a test
    based development approach when implementing the class and methods.

## Part B (10 marks)

Create a file `test-markovChain.R` with appropriate unit tests for the
class and methods from part A except for the `stationary_distribution`
and `visualise` methods. Write at least one test per method and one test
for the class itself.

## Part C (12 marks)

Create a folder `markovSim` using `devtools` that follows the standard R
package structure. Add the R source files from parts A and B in the
appropriate places. Take care of the following points:

-   All unit tests from part B should pass.
-   Make appropriate modifications of the DESCRIPTION file. Keep the
    details of the package brief.

Create an R package called `markovSim`. The result should be a file
named `markovSim_0.0.1.tar.gz` if you have chosen version number 0.0.1.

## Part D (15 marks)

Install and load the package `markovSim`. Use the code blocks below to
answer the following questions.

```{r}

install.packages("markovSim_0.0.1.tar.gz", repos = NULL, type = "source")
library(markovSim)

```

### Part D(i) (8 marks)

Consider the Markov chain starting from state 1 and with transition
matrix

$$
  P = \begin{pmatrix}
          0.6 & 0.4 \\
          0.3 & 0.7
      \end{pmatrix}
$$

Instantiate a `MarkovChain` object from the initial state and transition
matrix `P`. Perform the following operations and print out the results:

(i) Visualise the transition matrix using the `visualise` method.

```{r}
# provide your answer here

P <- matrix(c(0.6, 0.4,
              0.3, 0.7),
            nrow = 2, byrow = TRUE)

mc <- MarkovChain$new(P = P, initial_state = 1)

mc$visualise()

```

(ii) Simulate 20 paths of length 10.

```{r}
# set seed to ensure results are reproducable
set.seed(42)

# provide your answer here

# Simulate 20 paths of length 10
sim_res <- mc$simulate(n_steps = 10, n_paths = 20)

# Print the simulated paths
sim_res$paths
```

(iii) Compute the marginal distribution after 5 steps.

```{r}
# provide your answer here

pi_5 <- mc$marginal_distribution(t = 5)

pi_5
```

(iv) Compute the empirical marginal distributions along the 10 simulated
     paths.

```{r}
# provide your answer here
emp_marginals <- mc$empirical_marginal_distribution(sim_res$paths)

emp_marginals

```

(v) Compute an approximate stationary distribution.

```{r}
# provide your answer here

pi_sd <- mc$stationary_distribution()

pi_sd
```

(vi) Demonstrate the result stated in item (4) of the background section
     (convergence of relative frequencies for all states).

```{r}
# provide your answer here

N <- 200000L
sim_long <- mc$simulate(n_steps = N, n_paths = 1,)$paths[, 1]

x <- sim_long[-1L]

relfreq_1 <- cumsum(x == 1L) / seq_along(x)
relfreq_2 <- cumsum(x == 2L) / seq_along(x)

tail(cbind(relfreq_1, relfreq_2), 1)
pi_sd


check_N <- c(10L, 100L, 1e3L, 1e4L, 1e5L, 2e5L)
res <- cbind(
  N = check_N,
  freq_state_1 = relfreq_1[check_N],
  freq_state_2 = relfreq_2[check_N]
)



plot(relfreq_1, type = "l", xlab = "t", ylab = "frequency", ylim = c(0, 1))
abline(h = pi_sd[1], lty = 2)
lines(relfreq_2, type = "l")
abline(h = pi_sd["state_2"], lty = 2)

legend("top",
       legend = c("freq state 1", "freq state 2", "pi_1", "pi_2"),
       lty = c(1, 1, 2, 2), bty = "n")

#From values (code output) and graph we can see that the relative frequencies converge to the stationary distribution as N increases.

```

### Part D(ii) (7 marks)

Consider the transition matrix `P` defined in the following R snippet

```{r}
P <- matrix(c(
  0.40, 0.30, 0.20, 0.10, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
  0.10, 0.50, 0.20, 0.20, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
  0.20, 0.20, 0.40, 0.20, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
  0.30, 0.10, 0.20, 0.40, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00, 0.00,
  0.00, 0.00, 0.00, 0.00, 0.30, 0.20, 0.30, 0.20, 0.00, 0.00, 0.00, 0.00,
  0.00, 0.00, 0.00, 0.00, 0.20, 0.40, 0.20, 0.20, 0.00, 0.00, 0.00, 0.00,
  0.00, 0.00, 0.00, 0.00, 0.10, 0.30, 0.40, 0.20, 0.00, 0.00, 0.00, 0.00,
  0.00, 0.00, 0.00, 0.00, 0.20, 0.10, 0.30, 0.40, 0.00, 0.00, 0.00, 0.00,
  0.20, 0.10, 0.10, 0.10, 0.10, 0.10, 0.00, 0.00, 0.15, 0.15, 0.00, 0.00,
  0.10, 0.10, 0.00, 0.00, 0.10, 0.10, 0.10, 0.00, 0.00, 0.20, 0.15, 0.15,
  0.05, 0.05, 0.05, 0.05, 0.10, 0.10, 0.10, 0.10, 0.10, 0.00, 0.20, 0.10,
  0.00, 0.00, 0.10, 0.10, 0.05, 0.05, 0.05, 0.05, 0.20, 0.20, 0.10, 0.10
), nrow = 12, byrow = TRUE)
```

Use the class and methods from part A to show that a Markov chain with
this transition matrix cannot have a unique stationary distribution.

```{r}
# provide your anwser here

mc_1 = MarkovChain$new(P, 1)
mc_5 = MarkovChain$new(P, 5)

pi_1 = mc_1$stationary_distribution()
pi_5 = mc_5$stationary_distribution()

max(abs(pi_1 - as.numeric(pi_1 %*% P)))
max(abs(pi_5 - as.numeric(pi_5 %*% P)))

max(abs(pi_1 - pi_5))

#Both pi_1 = pi_1 * P and pi_5 = pi_5 * p, so both are stationary distributions
#however they are different distributions, so clearly there cannot be a stationary distribution

```

## Part E (15 marks)

Considering all the topics we have covered in the module, give an
example of something that, as a language, R does better than Python. It
could also be a feature which R has, but Python does not. There is more
than one correct answer, but you may provide *only one answer*. You may
*not* use examples which refer to topics in future modules such as "R is
better at regression" or "Python is better at deep learning". Refer to
topics or features you have covered in this module, but do not refer to
statistical or machine learning procedures covered in future modules.

Besides describing the example, you also need to provide accompanying
scripts in both R and Python that demonstrate the advantage described by
your example. A significant proportion of the marks is allocated to
providing the supporting example scripts and screenshots.

For your solution of this part of the question, create a folder `Q1E`
containing the following:

-   A text file `q1e.txt` where you describe your answer.
-   Example scripts `q1e.R` and `q1e.py` in R and Python, and any data
    files, to support your answer.
-   (optional) Screenshots in `.png`, `.jpg` or `.pdf` format if
    appropriate to your answer.
