---
title: \normalsize Assessment 3 \hfill MATH70094 Programming for Data Science \hfill Autumn 2025
    \rule{\linewidth}{0.5mm}
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NAME: GRAHAM DIB and CID: 01709270

# Question 2 (45 marks)

**The only non-base-R package you are allowed to load in this question
is testthat.**

Given a time series of length $n$ with data
$y_1,\dots,y_n\in\mathbb{R}$, we want to compare different ways for
forecasting future values $y_{k}$ of the time series at times $k>n$.
Depending on what statistical model we think has generated the data, a
different forecasting method is appropriate. Here, we consider two
statistical models:

(1) The time series was generated by $y_k = \epsilon_k$, where the
    $\epsilon_k$ are independent centred random variables. In this case,
    the time series has no trend and has expectation zero at all times.\
(2) The time series is an 'autoregressive process of order 1', where
    $y_{k+1} = \phi y_k + \epsilon_k$ for $k\in\mathbb{N}$ with a scalar
    $\phi$ and independent standard Gaussian random variables
    $\epsilon_k$ (more general autoregressive processes can be defined).
    The value of $\phi$ decides how the time series evolves over time.

Relative to the two models, we forecast future values of the time series
as follows:

(1) Using the rolling mean: Given a window length $h\in\mathbb{N}$,
    compute the 'rolling mean'
    $\bar{y}_h = h^{-1}(y_{n-h+1} + y_{n-h+2} + \dots + y_n)$. The
    predicted value of the time series at time $k>n$ is $\bar{y}_h$ for
    any $k$.
(2) The parameter $\phi$ can be estimated by
    $\hat{\phi} = \frac{\sum_{k=2}^n y_{k-1} y_k}{\sum_{k=2}^n (y_{k-1})^2}$.
    The predicted value of the time series at time $n+1$ is
    $\hat y_{n+1} = \hat\phi y_n$, the value at $n+2$ is
    $\hat y_{n+2} = \hat\phi \hat y_{n+1}$ etc.

Your task in this question is to write three different S3 classes in R
specified as follows:

1.  A class `TimeSeriesModel` with two methods `forecast` and
    `plot_forecast`. The two methods have no implementations, but both
    take as argument an integer `window_prediction_length` to indicate
    the number of future time steps that should be forecast. This class
    serves as a common parent class for the subclasses
    `RollingMeanModel` and `AR1Model` below. The constructor should not
    do anything as we will never instantiate objects from this class, so
    you can just use

```{r}
TimeSeriesModel <- function() {
  # TimeSeriesModel is abstract so use a subclass
}
```

2.  A subclass `RollingMeanModel` of `TimeSeriesModel` with

-   a constructor method that instantiates an object of
    `RollingMeanModel` given a time series (a `ts` object, check `?ts`)
    and an integer argument `window_train_length`; the latter is used to
    set the attribute `window_mean` that records the average value of
    the time series over the last `window_train_length` values,
-   a forecast method that overrides the forecast method from
    `TimeSeriesModel` and predicts the value of the time series for
    `window_prediction_length` future steps by `window_mean`,
-   a plot method `plot_forecast` that plots the time series model using
    the `plot.ts` function as well as the values of a 5-step prediction
    in the same graph.

3.  A subclass `AR1Model` of `TimeSeriesModel` with

-   a constructor method that instantiates an object of `AR1Model` given
    a time series; an object is instantiated by estimating the phi
    coefficient using the estimator $\hat\phi$ from above; the estimator
    is used to set an attribute of the same name,
-   a forecast method that overrides the forecast method from
    `TimeSeriesModel` and predicts the value of the time series for
    `window_prediction_length` future steps as follows: at time one in
    the future the prediction is $\hat y_{n+1}=\hat\phi*y_{n}$ with
    $y_{n}$ the last value of the training data, at time two by
    $\hat y_{n+2}=\hat\phi*\hat y_{n+1}$, etc,
-   a plot method `plot_forecast` that plots the time series model using
    the `plot.ts` function as well as the values of a 5-step prediction.

Below, you will be asked to write and use these classes in a test driven
development approach.

## Code clarity (3 mark)

There is a famous saying among software developers that code is read
more often than it is written. Marks will be awarded (or not awarded)
based on the clarity of the code and appropriate use of comments.

## Part A (10 marks)

Provide an interface (only class names, method names and empty method
bodies, S3-generics) for the three classes as specified above. Every
class and every method should have a short comment describing the
expected functionality. Do not provide implementations for the methods.

```{r}
# write the interfaces here


# TimeSeriesModel constructor 
TimeSeriesModel <- function() {
}


# Generic forecast method for all time series models
forecast <- function(object, window_prediction_length, ...) {
  UseMethod("forecast")
}

# Generic plotting method for forecasting
plot_forecast <- function(object, window_prediction_length, ...) {
  UseMethod("plot_forecast")
}


# Abstract forecast method 
forecast.TimeSeriesModel <- function(object, window_prediction_length, ...) {
}

# Abstract plot_forecast method
plot_forecast.TimeSeriesModel <- function(object, window_prediction_length, ...) {
}


# Constructor for RollingMeanModel
RollingMeanModel <- function(ts_object, window_train_length) {
}

# forecast method for RollingMeanModel
# Should return future predictions equal to the stored rolling mean
forecast.RollingMeanModel <- function(object, window_prediction_length, ...) {
}

# plot_forecast method for RollingMeanModel
# Should plot the ts object and overlay 5-step rolling-mean predictions
plot_forecast.RollingMeanModel <- function(object, window_prediction_length, ...) {
}


# Constructor for AR1Model
# Should estimate phi and store it along with the series
AR1Model <- function(ts_object) {
}

# forecast method for AR1Model
# Should iteratively forecast using phi * previous_value
forecast.AR1Model <- function(object, window_prediction_length, ...) {
}

# plot_forecast method for AR1Model
# Should plot the ts object and overlay 5-step AR(1) predictions
plot_forecast.AR1Model <- function(object, window_prediction_length, ...) {
}
```

## Part B (10 marks)

Use the `testthat` library (you may have to install it first) to write
two test functions with `test_that` to test if your class codes work as
expected:

1.  For use during testing, create a time series object by defining a
    vector repeating the value 1 for 50 times and applying the `ts`
    function. Generate from this a `RollingMeanModel` object with
    `window_train_length` equal to 5, and an `AR1Model` object.
2.  A test function that checks if the models have been constructed
    correctly. There is some flexibility here. You could test for
    example if the internal variables have been set properly.
3.  A test function that checks if the model predictions have the
    correct length (as specified by `window_prediction_length`).

All tests should \emph{fail} at this point. Include `error=TRUE` in the
RMarkdown cell to prevent failing tests from stopping code execution.

```{r, error=TRUE}
library(testthat)

#1.

# Time series of 50 ones
ts_data <- ts(rep(1, 50))

# RollingMeanModel with window_train_length = 5
rm <- RollingMeanModel(ts_data, window_train_length = 5)

# AR1Model
ar1 <- AR1Model(ts_data)

#2.

test_that("Constructors", {
  # RollingMeanModel should have correct class
  expect_s3_class(rm, "RollingMeanModel")
  expect_s3_class(rm, "TimeSeriesModel")

  # AR1Model should have correct class
  expect_s3_class(ar1, "AR1Model")
  expect_s3_class(ar1, "TimeSeriesModel")

  # check if window_mean is stored correctly.
  expect_equal(attr(rm, "window_mean"), 1)

  # check phi stored correctly
  expect_equal(attr(ar1, "phi"), 1)
})

#3.

test_that("Forecasts", {
  window_prediction_length <- 10

  expect_error({
    rm_forecast  <- forecast(rm, window_prediction_length)
    ar1_forecast <- forecast(ar1, window_prediction_length)
  
  
    expect_length(rm_forecast,  window_prediction_length)
    expect_length(ar1_forecast, window_prediction_length)
  }, NA)
})
```

## Part C (10 marks)

Provide implementations for the three classes from part A. All tests
from part B should now pass without errors or failures.

```{r}
# write the interfaces here


# TimeSeriesModel constructor 
TimeSeriesModel <- function() {
}


# Generic forecast method for all time series models
forecast <- function(object, window_prediction_length, ...) {
  UseMethod("forecast")
}

# Generic plotting method for forecasting
plot_forecast <- function(object, window_prediction_length, ...) {
  UseMethod("plot_forecast")
}


# Abstract forecast method 
forecast.TimeSeriesModel <- function(object, window_prediction_length, ...) {
}

# Abstract plot_forecast method
plot_forecast.TimeSeriesModel <- function(object, window_prediction_length, ...) {
}


# Constructor for RollingMeanModel
RollingMeanModel <- function(ts_object, window_train_length) {
  
    
  window_mean <- mean(tail(ts_object, window_train_length))

  obj <- list(
    ts_object = ts_object,
    window_train_length = window_train_length
  )

  attr(obj, "window_mean") <- window_mean

  class(obj) <- c("RollingMeanModel", "TimeSeriesModel")
  obj
}

# forecast method for RollingMeanModel
# Should return future predictions equal to the stored rolling mean
forecast.RollingMeanModel <- function(object, window_prediction_length, ...) {
  
  window_mean <- attr(object, "window_mean")
  rep(window_mean, window_prediction_length)

}

# plot_forecast method for RollingMeanModel
# Should plot the ts object and overlay 5-step rolling-mean predictions
plot_forecast.RollingMeanModel <- function(object, window_prediction_length, ...) {
  
  ts_object <- object$ts_object

  plot.ts(ts_object,
        main = "RollingMeanModel",
        ylab = "Value",
        xlab = "Time")

  forecast <- forecast(object, window_prediction_length)
  
  prev <- tail(time(ts_object), 1)
  freq <- frequency(ts_object)
  new <- prev + seq_len(window_prediction_length) / freq

  lines(new, forecast)
  points(new, forecast)

}


# Constructor for AR1Model
# Should estimate phi and store it along with the series
AR1Model <- function(ts_object) {
  y <- as.numeric(ts_object)
  n <- length(y)

  # Estimate phi using eq
  num <- sum(y[-1L] * y[-n])
  den <- sum(y[-n]^2)
  if (den == 0) {
    phi_hat <- 0
  } else {
    phi_hat <- num / den
  }

  obj <- list(
    ts_object = ts_object
  )
  
  attr(obj, "phi") <- phi_hat

  class(obj) <- c("AR1Model", "TimeSeriesModel")
  obj


}

# forecast method for AR1Model
# Should iteratively forecast using phi * previous_value
forecast.AR1Model <- function(object, window_prediction_length, ...) {
  
  ts_object <- object$ts_object

  
  phi <- attr(object, "phi")
  y_prev <- tail(as.numeric(ts_object), 1)
  
  forecast <- numeric(window_prediction_length)
  if (window_prediction_length >= 1L) {
    forecast[1] <- phi * y_prev
    if (window_prediction_length >= 2L) {
      for (k in 2:window_prediction_length) {
        forecast[k] <- phi * forecast[k - 1L]
      }
    }
  }
  forecast
}

# plot_forecast method for AR1Model
# Should plot the ts object and overlay 5-step AR(1) predictions
plot_forecast.AR1Model <- function(object, window_prediction_length, ...) {
  
  ts_object <- object$ts_object

  plot.ts(ts_object,
        main = "AR1Model: data and forecast",
        ylab = "Value",
        xlab = "Time")
  
  forecast <- forecast(object, window_prediction_length)

  prev <- tail(time(ts_object), 1)
  freq <- frequency(ts_object)
  new <- prev + seq_len(window_prediction_length) / freq

  lines(new, forecast)
  points(new, forecast)


}
```

## Part D (12 marks)

### Part D(i) (8 marks)

Add a function `compare_models` (not part of any class) that takes as
arguments a time series `train`, a time series `test` and a
`window_train_length`. Within the function, instantiate from the `train`
time series a `RollingMeanModel` object and an `AR1Model` object. The
`compare_models` function should then return TRUE if the mean square
error between `test` and the forecast of the same length by the
`RollingMeanModel` object is smaller than the mean square error between
`test` and the forecast of the same length by the `AR1Model` object.

Write also a test function (again using `test_that`) that checks if the
`compare_models` function works as expected (and ideally write the test
method first). The test should pass.

```{r}
# add your solution here
compare_models <- function(train, test, window_train_length) {

  
  rm  <- RollingMeanModel(train, window_train_length)
  ar1 <- AR1Model(train)
  
  forecast_length <- length(test)
  rm_forecast  <- forecast(rm,  forecast_length)
  ar1_forecast <- forecast(ar1, forecast_length)
  
  test_vals <- as.numeric(test)
  mse_rm  <- mean((test_vals - rm_forecast)^2)
  mse_ar1 <- mean((test_vals - ar1_forecast)^2)

  
  mse_rm < mse_ar1

}

test_that("Compare_models check", {
  
  train <- ts(c(1,1,1,1,2,2))
  test <- ts(rep(1,5))
  
  res <- compare_models(train, test, window_train_length = length(train))
  
  
  expect_true(res)
})



```

### Part D(ii) (4 marks)

Read in the data from the file `time_series_data.csv`. Split this data
into two data sets and create two time series objects `train` and `test`
using the `ts` function, where `train` consists of the first 150 data
points and `test` of the remaining data. Using the `compare_models`
functions from Part D(i), which model (out of `RollingMeanModel` and
`AR1Model`) performs better in terms of forecasting for this data set?

```{r}
# add your solution here

data <- read.csv("time_series_data.csv")

y <- as.numeric(data[[1]])

train <- ts(y[1:150])
test  <- ts(y[151:length(y)])

res <- compare_models(train, test, window_train_length = 5)
res

if (res) {
  cat("RollingMeanModel performs better i.e RollingMeanModel has smaller MSE).\n")
} else {
  cat("AR1Model performs better i.e AR1Model has smaller MSE.\n")
}
```
